# Descriptive analysis {.tabset .tabset-fade}

<!-- ======================================================= -->

## Overview {.tabset .tabset-fade .tabset-pills}

This tab demonstrates the use of **gtsummary** and **dplyr** to produce descriptive statistics.

1.  Browse data: get a quick overview of your dataset using the **skimr** package

2.  Summary statistics: mean, median, range, standard deviations, percentiles

3.  Frequency / cross-tabs: counts and proportions

4.  Statistical tests: t-tests, wilcoxon rank sum, kruskal-wallis and chi-squares

5.  Correlations

<!-- ======================================================= -->

## Preparation {.tabset .tabset-fade .tabset-pills}

<h2>

Preparation

</h2>

### Packages

This code chunk shows the loading of packages required for the analyses.

```{r}
pacman::p_load(rio,          # File import
               here,         # File locator
               skimr,        # get overview of data
               tidyverse,    # data management + ggplot2 graphics, 
               gtsummary,    # summary statistics and tests 
               corrr         # correlation analayis for numeric variables
               )
```

### Load data

The example dataset used in this section:

-   Linelist of individual cases from a simulated epidemic

The dataset is imported using the `import()` function from the *rio* package. See the *page on importing data* for various ways to import data.

```{r, echo=F}
# import the linelist into R
linelist <- rio::import(here::here("data", "linelist_cleaned.rds"))
```

```{r, eval=F}
# import the linelist
linelist <- rio::import("linelist_cleaned.xlsx")
```

The first 50 rows of the linelist are displayed below.

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist, 50), rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
```

### Clean data

```{r}

## make sure that age variable is numeric 
linelist <- linelist %>% 
  mutate(age = as.numeric(age))
```

<!-- ======================================================= -->

## Browse data {.tabset .tabset-fade .tabset-pills}

<h2>

Browse data

</h2>

### Base R

You can use the `summary` function to get information about variables and data sets.

For a numeric variable it will give you the minimum, median, mean and max as well as the 1st quartile (= 25th percentile) and the 3rd quartile (= 75th percentile)

```{r descriptive_browse}

## get information about a numeric variable 
summary(linelist$age)

```

You can also get an overview of each variable in a whole dataset.

```{r descriptive_browse_df}

## get information about each variable in a dataset 
summary(linelist)
```

### `skimr` package

Using the `skimr` package you can get a more detailed overview of each of the variables in your dataset.

```{r descriptive_browse_skimr}

## get information about each variable in a dataset 
skim(linelist)
```

<!-- ======================================================= -->

## Summary Statistics {#summarystats .tabset .tabset-fade .tabset-pills}

### `gtsummary` package

Using `gtsummary` you can create a table with different summary statistics, for example mean, median, range, standard deviation and percentiles. You can also show these all in one table.

#### **Mean**

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column.

```{r descriptive_gt_mean}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with mean
  tbl_summary(statistic = age ~ "{mean}")

```

#### **Median**

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r descriptive_gt_median}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with median
  tbl_summary(statistic = age ~ "{median}")

```

#### **Range**

The range here is the minimum and maximum values for the variable. (see *percentiles* for interquartile range) Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r descriptive_gt_range}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with range 
  tbl_summary(statistic = age ~ "{min}, {max}")

```

#### **Standard deviation**

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r descriptive_gt_sd}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with standard deviation
  tbl_summary(statistic = age ~ "{sd}")

```

#### **Percentile**

To return percentiles you can type in one value that you would like, or you can type in multiple (e.g. to return the interquartile range).

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r descriptive_gt_percentile}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with interquartile range 
  tbl_summary(statistic = age ~ "{p25}, {p75}")

```

#### **Combined table**

You can combine all of the previously shown elements in one table by choosing which statistics you want to show. To do this you need to tell the function that you want to get a table back by entering the type as "continuous2".

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r descriptive_gt_combined}

linelist %>% 
  ## only keep variable of interest
  select(age) %>% 
  ## create summary table with interquartile range 
  tbl_summary(
    ## tell the function you want to get multiple statistics back 
    type = age ~ "continuous2",
    ## define which statistics you want to get back 
    statistic = age ~ c(
    "{mean} ({sd})", 
    "{median} ({p25}, {p75})",
    "{min}, {max}")
    )

```

### `dplyr` package

You can also use `dplyr` to create a table with different summary statistics, for example mean, median, range, standard deviation and percentiles. You can also show these all in one table. The difference with using dplyr is that the output is not automatically formatted as nicely as with `gtsummary`

#### **Mean**

Note the argument `na.rm = TRUE`, which removes missing values from the calculation.\
If missing values are not excluded, the returned value will be NA (missing).

```{r descriptive_dplyr_mean}

linelist %>% 
  ## get the mean value of age while excluding missings
  summarise(mean = mean(age, na.rm = TRUE))
```

#### **Median**

Note the argument `na.rm = TRUE`, which removes missing values from the calculation.\
If missing values are not excluded, the returned value will be NA (missing).

```{r descriptive_dplyr_median}

linelist %>% 
  ## get the median value of age while excluding missings
  summarise(median = median(age, na.rm = TRUE))
```

#### **Range**

Note the argument `na.rm = TRUE`, which removes missing values from the calculation.\
If missing values are not excluded, the returned value will be NA (missing).

```{r descriptive_dplyr_range}

linelist %>% 
  ## get the range value of age while excluding missings
  summarise(range = range(age, na.rm = TRUE))
```

#### **Standard Deviation**

Note the argument `na.rm = TRUE`, which removes missing values from the calculation.\
If missing values are not excluded, the returned value will be NA (missing).

```{r descriptive_dplyr_sd}

linelist %>% 
  ## get the range value of age while excluding missings
  summarise(sd = sd(age, na.rm = TRUE))

```

#### **Percentile**
To return percentiles you can type in one value that you would like, or you can type in multiple (e.g. to return the interquartile range).

Note the argument `na.rm = TRUE`, which removes missing values from the calculation.\
If missing values are not excluded, the returned value will be NA (missing).

```{r descriptive_dplyr_percentile}


linelist %>% 
  ## get the default percentile values of age while excluding missings 
  ## these are 0%,  25%,  50%,  75%, 100%
  summarise(percentiles = quantile(age, na.rm = TRUE))


linelist %>% 
  ## get specified percentile values of age while excluding missings 
  ## these are 0%, 50%,  75%, 98%
  summarise(percentiles = quantile(age,
                                   probs = c(.05, 0.5, 0.75, 0.98), 
                                   na.rm=TRUE))
```


#### **Combined table** 

You can combine all of the previously shown elements in one table by choosing 
which statistics you want to show. In dplyr you will need to use the `str_c` 
function from **stringr** to combine outputs for the IQR and the range in to one 
cell, separated by a comma. 

Note that this automatically excludes all missing values. If missing values are not excluded, the returned value will be NA (missing). The number of missing values is seen in the Unknown column).

```{r}

linelist %>% 
  summarise(
    ## get the mean 
    mean = mean(age, na.rm = TRUE),
    ## get the standard deviation
    SD = sd(age, na.rm = TRUE),
    ## get the median 
    median = median(age, na.rm = TRUE), 
    ## collapse the IQR separated by a comma
    IQR = str_c(
      quantile(age, probs = c(0.25, 0.75), na.rm = TRUE), 
      collapse = ", "
      ), 
    ## collapse the range separated by a comma
    Range = str_c(
      range(age, na.rm = TRUE), 
      collapse = ", "
    )
  )

```



<!-- ======================================================= -->

## Frequency/cross-tabs {.tabset .tabset-fade .tabset-pills}

### `gtsummary` package

TODO: Note that percentages are calculated without missings 

Using `gtsummary` you can create a table with different counts and proportions 
for variables with two or more categories, as well as grouping by another variable. 

#### **One way table**
To produce the counts of a single variable we can use the `tbl_summary` function. 
Note that here, the fever variable is yes/no (dichotomous) and `tbl_summary`
automatically only presents the "yes" row. 
To show all levels you could use the type argument to choose categorical, 
e.g. `tbl_summary(type = fever ~ "categorical")`. 
```{r descriptive_gt_freq_one}

linelist %>% 
  ## only keep the variable interested in
  select(fever) %>% 
  ## produce summary table
  tbl_summary()
```

#### **Multiple variable one way table**

You can also show multiple variables below each other simply by adding them to 
`select`. 

```{r descriptive_gt_freq_one_mult}

linelist %>% 
  ## only keep the variable interested in
  select(fever, gender) %>% 
  ## produce summary table
  tbl_summary()

```

#### **Two way table**

There are two options to produce a two-by-two table (i.e. comparing two variables). 
One option is to use `tbl_cross`, however this function only accepts two variables
at once. The option below with `tbl_summary` allows more variables. 

```{r descriptive_gt_freq_two}

linelist %>% 
  ## only keep the variable interested in
  select(fever, outcome, gender) %>% 
  ## produce summary table stratified by gender
  tbl_summary(by = gender) %>% 
  ## add a column for the totals
  add_overall()

```

#### **Three way table**
Producing counts based on three variables (adding a stratifier). 

```{r descriptive_gt_freq_three, eval=F}
## TODO: add stratified tables when available 

# table_3vars <- table(linelist$fever, linelist$gender, linelist$outcome)
# 
# ftable(table_3vars)
```

### `dplyr` package

Creating cross tabulations with dplyr is less straightforward, as this does not
fit within the **tidyverse** dataset structure. It is still useful to demonstrate
though as the data produced can be used for plotting *reference ggplot section*. 
Another option is to use the **janitor** package [`tabyl` function](http://sfirke.github.io/janitor/reference/tabyl.html). 

#### **One way table**
Producing counts and proportions for a single variable. 
To see how to do this for multiple variables - *reference for-loop section*. 

```{r descriptive_dplyr_freq_one}

linelist %>% 
  ## count the variable of interest
  count(fever) %>% 
  ## calculate proportion 
  mutate(percentage = n / sum(n) * 100)
```

#### **Two way table**

Producing counts and proportions based on a grouping variable. Here we use the 
**dplyr** `group_by` function, for more information see the 
*reference grouping and aggregating section*. 
You can calculate the percentages of the total by using `ungroup()` after `count(...)`. 

Note that it is possible to change the bellow table to wide format, making it 
more like a two-by-two (cross tabulation), using the **tidyr** `pivot_wider` function.
This would be done by adding this to the end of the code blow: 
`pivot_wider(names_from = gender, values_from = c(n, percentage))`
For more information see the *reference pivot section*. 

```{r descriptive_dplyr_freq_two}

linelist %>% 
  ## do everything by gender 
  group_by(gender) %>% 
  ## count the variable of interest
  count(fever) %>% 
  ## calculate proportion 
  ## note that the denominator here is the sum of each gender
  mutate(percentage = n / sum(n) * 100)

```

#### **Three way table**

Producing counts based on three variables (adding a stratifier). 

```{r descriptive_dplyr_freq_three}

linelist %>% 
  ## do everything by gender and outcome 
  group_by(gender, outcome) %>% 
  ## count the variable of interest 
  count(fever) %>% 
  ## calculate the proportion
  ## note that the denominator here is the sum of each group combination
  mutate(percentage = n / sum(n) * 100)

```


<!-- ======================================================= -->

## Statistical tests {.tabset .tabset-fade .tabset-pills}

### `gtsummary` package

Performing statistical tests of comparison with `tbl_summary` is done by using 
`add_p` function and specifying which test to use. 
It is possible to get p-values corrected for multiple testing by using the
`add_q` function. 

#### T-tests 

Compare the difference in means for a continuous variable in two groups. 
For example compare the mean age by patient outcome. 

```{r ttest_gt}

linelist %>% 
  ## only keep variables of interested
  select(age, outcome) %>% 
  ## produce summary table
  tbl_summary(
    ## specify what statistic want to show
    statistic = age ~ "{mean} ({sd})", 
    ## specify the grouping variable
    by = outcome) %>% 
  ## specify what test want to perform
  add_p(age ~ "t.test")

```

#### Wilcoxon rank sum test

Compare the distribution of a continuous variable in two groups. The default 
is to use the Wilcoxon rank sum test and the median (IQR) when comparing two 
groups. However for non-normally distributed data or comparing multiple groups, 
the Kruskal-wallis test is more appropriate. 

```{r wilcox_gt}

linelist %>% 
  ## only keep variables of interested
  select(age, outcome) %>% 
  ## produce summary table
  tbl_summary(
    ## specify what statistic want to show (default so could remove)
    statistic = age ~ "{median} ({p25}, {p75})", 
    ## specify the grouping variable
    by = outcome) %>% 
  ## specify what test want to perform (default so could leave brackets empty)
  add_p(age ~ "wilcox.test")

```

#### Kruskal-wallis test

Compare the distribution of a continuous variable in two or more groups, 
regardless of whether the data is normally distributed. 

```{r kruskal_gt}

linelist %>% 
  ## only keep variables of interested
  select(age, outcome) %>% 
  ## produce summary table
  tbl_summary(
    ## specify what statistic want to show (default so could remove)
    statistic = age ~ "{median} ({p25}, {p75})", 
    ## specify the grouping variable
    by = outcome) %>% 
  ## specify what test want to perform
  add_p(age ~ "kruskal.test")

```

#### Chi-squared test 

Compare the proportions of a categorical variable in two groups. The default is 
to perform a chi-squared test of independence with continuity correction, but if 
any expected call count is below 5 then a Fisher's exact test is used. 

```{r chi_gt}

linelist %>% 
  ## only keep variables of interested
  select(gender, outcome) %>% 
  ## produce summary table
  tbl_summary(
    ## specify the grouping variable
    by = outcome
  ) %>% 
  ## specify what test want to perform
  add_p()

```

### `dplyr` package

Performing statistical tests in `dplyr` alone is very dense, again because it 
does not fit within the tidy-data framework. It requires using `purrr` to create
a list of dataframes for each of the subgroups you want to compare. 
An easier alternative may be the `rstatix` package. 

#### T-tests 

```{r ttest_dplyr}

linelist %>% 
  ## only keep variables of interest
  select(age, outcome) %>% 
  ## drop those missing outcome 
  filter(!is.na(outcome)) %>% 
  ## specify the grouping variable
  group_by(outcome) %>% 
  ## create a subset of data for each group (as a list)
  nest() %>% 
  ## spread in to wide format
  pivot_wider(names_from = outcome, values_from = data) %>% 
  mutate(
    ## calculate the mean age for the death group
    Death_mean = map(Death, ~mean(.x$age, na.rm = TRUE)),
    ## calculate the sd among dead 
    Death_sd = map(Death, ~sd(.x$age, na.rm = TRUE)),
    ## calculate the mean age for the recover group
    Recover_mean = map(Recover, ~mean(.x$age, na.rm = TRUE)), 
    ## calculate the sd among recovered 
    Recover_sd = map(Recover, ~sd(.x$age, na.rm = TRUE)),
    ## using both grouped data sets compare mean age with a t-test
    ## keep only the p.value
    t_test = map2(Death, Recover, ~t.test(.x$age, .y$age)$p.value)
  ) %>% 
  ## drop datasets 
  select(-Death, -Recover) %>% 
  ## return a dataset with the medians and p.value (drop missing)
  unnest(cols = everything())

```


#### Wilcoxon rank sum test

```{r wilcox_dplyr}

linelist %>% 
  ## only keep variables of interest
  select(age, outcome) %>% 
  ## drop those missing outcome 
  filter(!is.na(outcome)) %>% 
  ## specify the grouping variable
  group_by(outcome) %>% 
  ## create a subset of data for each group (as a list)
  nest() %>% 
  ## spread in to wide format
  pivot_wider(names_from = outcome, values_from = data) %>% 
  mutate(
    ## calculate the median age for the death group
    Death_median = map(Death, ~median(.x$age, na.rm = TRUE)),
    ## calculate the sd among dead 
    Death_iqr = map(Death, ~str_c(
      quantile(.x$age, probs = c(0.25, 0.75), na.rm = TRUE), 
      collapse = ", "
      )),
    ## calculate the median age for the recover group
    Recover_median = map(Recover, ~median(.x$age, na.rm = TRUE)), 
    ## calculate the sd among recovered 
    Recover_iqr = map(Recover, ~str_c(
      quantile(.x$age, probs = c(0.25, 0.75), na.rm = TRUE), 
      collapse = ", "
      )),
    ## using both grouped data sets compare age distribution with a wilcox test
    ## keep only the p.value
    wilcox = map2(Death, Recover, ~wilcox.test(.x$age, .y$age)$p.value)
  ) %>% 
  ## drop datasets 
  select(-Death, -Recover) %>% 
  ## return a dataset with the medians and p.value (drop missing)
  unnest(cols = everything())

```

#### Kruskal-wallis test


```{r kruskal_dplyr}

linelist %>% 
  ## only keep variables of interest
  select(age, outcome) %>% 
  ## drop those missing outcome 
  filter(!is.na(outcome)) %>% 
  ## specify the grouping variable
  group_by(outcome) %>% 
  ## create a subset of data for each group (as a list)
  nest() %>% 
  ## spread in to wide format
  pivot_wider(names_from = outcome, values_from = data) %>% 
  mutate(
    ## calculate the median age for the death group
    Death_median = map(Death, ~median(.x$age, na.rm = TRUE)),
    ## calculate the sd among dead 
    Death_iqr = map(Death, ~str_c(
      quantile(.x$age, probs = c(0.25, 0.75), na.rm = TRUE), 
      collapse = ", "
      )),
    ## calculate the median age for the recover group
    Recover_median = map(Recover, ~median(.x$age, na.rm = TRUE)), 
    ## calculate the sd among recovered 
    Recover_iqr = map(Recover, ~str_c(
      quantile(.x$age, probs = c(0.25, 0.75), na.rm = TRUE), 
      collapse = ", "
      )),
    ## using the original data set compare age distribution with a kruskal test
    ## keep only the p.value
    kruskal = kruskal.test(linelist$age, linelist$outcome)$p.value
  ) %>% 
  ## drop datasets 
  select(-Death, -Recover) %>% 
  ## return a dataset with the medians and p.value (drop missing)
  unnest(cols = everything())

```

#### Chi-squared test 


```{r}
linelist %>% 
  ## do everything by gender 
  group_by(outcome) %>% 
  ## count the variable of interest
  count(gender) %>% 
  ## calculate proportion 
  ## note that the denominator here is the sum of each gender
  mutate(percentage = n / sum(n) * 100) %>% 
  pivot_wider(names_from = outcome, values_from = c(n, percentage)) %>% 
  filter(!is.na(gender)) %>% 
  mutate(pval = chisq.test(linelist$gender, linelist$outcome)$p.value)
```


### `base` package

You can also just use the `base` functions to produce the results of statistical 
tests. The outputs of these are however usually lists, and so are harder to 
manipulate. 

#### T-tests 

```{r ttest_base}

## compare mean age by outcome group with a t-test
t.test(age ~ outcome, data = linelist)
```


#### Wilcoxon rank sum test

```{r wilcox_base}

## compare age distribution by outcome group with a wilcox test
wilcox.test(age ~ outcome, data = linelist)

```

#### Kruskal-wallis test


```{r }

## compare age distribution by outcome group with a kruskal-wallis test
kruskal.test(age ~ outcome, linelist)

```

#### Chi-squared test 


```{r}

## compare the proportions in each group with a chi-squared test
chisq.test(linelist$gender, linelist$outcome)

```




<!-- ======================================================= -->

## Correlations {.tabset .tabset-fade .tabset-pills}

Correlation between numeric variables can be investigated using the tidyverse 
`corrr` package. It allows you to compute correlations using Pearson, Kendall
tau or Spearman rho. The package creates a table and also has a function to 
automatically plot the values. 
```{r}

correlation_tab <- linelist %>% 
  ## pick the numeric variables of interest
  select(generation, age, ct_blood, days_onset_hosp, wt_kg, ht_cm) %>% 
  ## create correlation table (using default pearson)
  correlate()

## remove duplicate entries (the table is mirrored) 
correlation_tab <- correlation_tab %>% 
  shave()


## view correlation table 
correlation_tab

## plot correlations 
rplot(correlation_tab)
```


<!-- ======================================================= -->

## Resources {.tabset .tabset-fade .tabset-pills}

Much of the information in this page is adapted from these resources and vignettes online:  

[gtsummary](http://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html)
[dplyr](https://dplyr.tidyverse.org/articles/grouping.html)
[corrr](https://corrr.tidymodels.org/articles/using-corrr.html)
[sthda correlation](http://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r)
