# Style and editorial notes { }


## Style  


### Text style


**Package and function names**  

Package names are written in bold (e.g. **dplyr**) and functions are written like this: `mutate()`. Packages referenced either in text or within code like this: `dplyr::mutate()`  



**Types of notes**  

<span style="color: black;">**_NOTE:_** This is a note</span>

<span style="color: darkgreen;">**_TIP:_** This is a tip.</span>

<span style="color: orange;">**_CAUTION:_** This is a cautionary note.</span>

<span style="color: red;">**_DANGER:_** This is a warning.</span>



### **tidyverse**  

This handbook generally uses **tidyverse** R coding style. Read more [here](https://www.tidyverse.org/)



### Code readability  

We chose to frequently write code on new lines, in order to offer more understandable comments. As a result, code that *could* be written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%                    # group the rows by 'name'
  slice_max(date, n = 1, with_ties = F) # if there's a tie (of date), take the first row
```

...is often written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%   # group the rows by 'name'
  slice_max(
    date,              # keep row per group with maximum date value 
    n = 1,             # keep only the single highest row 
    with_ties = F)     # if there's a tie (of date), take the first row
```




## Editorial decisions  

Below, we track significant editorial decisions around package and function choice. If you disagree or want to offer a new tool, please join/start a conversation on our Github page.    


**Table of package, function, and other editorial decisions**  



Subject           |     Considered      |   Outcome & date    |    Brief rationale   
----------------- | --------------------|---------------------|--------------------------------------------------    
Epiweeks          | aweek, lubridate    | lubridate, Dec 2020 | consistency, package maintenance prospects  
   




