# Style and editorial notes { }


## Style  


### Notes {-}  

Here are the types of notes you may encounter in the handbook:  

<span style="color: black;">**_NOTE:_** This is a note</span>

<span style="color: darkgreen;">**_TIP:_** This is a tip.</span>

<span style="color: orange;">**_CAUTION:_** This is a cautionary note.</span>

<span style="color: red;">**_DANGER:_** This is a warning.</span>


### Package choice {-}

This handbook generally prioritizes R packages and functions from the **tidyverse**. Tidyverse is a collection of R packages designed for data science that share underlying grammar and data structures. All tidyverse packages can be installed or loaded via the **tidyverse** package. Read more at the [tidyverse website](https://www.tidyverse.org/).  

When applicable, we also offer code options using **base** R - the packages and functions that come with R at installation. This is because we recognize that some in this book's audience may not have reliable internet to download many packages.  


### Coding style {-}

In the handbook, we frequently utilize new lines in our code, making it appear "long". We do this to offer more comments, make it easier to read on a narrow screen (no sideways scrolling needed), and to be clear which arguments belong to which function. Generally, R is not affected by such new lines.  

As a result, code that *could* be written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%                    # group the rows by 'name'
  slice_max(date, n = 1, with_ties = F) # if there's a tie (of date), take the first row
```

...is written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%   # group the rows by 'name'
  slice_max(
    date,              # keep row per group with maximum date value 
    n = 1,             # keep only the single highest row 
    with_ties = F)     # if there's a tie (of date), take the first row
```


In the narrative text, package names are written in bold (e.g. **dplyr**) and functions are written like this: `mutate()`. We strive to be explicit about which package a function comes from, either by referencing this in nearby text or specifying the package explicitly in the code like this: `dplyr::mutate()`. See the page on [R basics] to learn more about packages and functions.  



## Editorial decisions  

Below, we track significant editorial decisions around package and function choice. If you disagree or want to offer a new tool, please join/start a conversation on our Github page.    


**Table of package, function, and other editorial decisions**  



Subject           |     Considered      |   Outcome & date    |    Brief rationale   
----------------- | --------------------|---------------------|--------------------------------------------------    
Epiweeks          | aweek, lubridate    | lubridate, Dec 2020 | consistency, package maintenance prospects  
ggplot labels     | labs(), ggtitle(), ylab(), xlab() | labs(), Feb 2021 | All labels in one function/one place |
   




