
# Epidemic modeling {#epidemicmodels .tabset .tabset-fade}  


<!-- ======================================================= -->
## Overview {.tabset .tabset-fade .tabset-pills}

There exists a growing body of tools for epidemic modelling that let us conduct
fairly complex analyses with minimal effort. This section will provide an
overview on how to use these tools to:

* estimate the effective reproduction number R<sub>t</sub> and related statistics
  such as the doubling time
* produce short-term projections of future incidence

It is *not* intended as an overview of the methodologies and statistical methods
underlying these tools, so please refer to the Resources tab for links to some
great web pages covering this. Make sure you have an understanding of
the methods before using these tools; this will ensure you can accurately
interpret their outputs.

Below is an example of one of the outputs we'll be producing in this section.

```{r out.width=c('25%', '25%'), fig.show='hold', echo=F}

## install and load packages
pacman::p_load(tidyverse, EpiNow2, EpiEstim, here, epicontacts, rio, projections)

## load linelist
linelist <- rio::import(here::here("data", "linelist_cleaned.rds"))

## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)

## ## estimate gamma generation time
## generation_time <- bootstrapped_dist_fit(
##   get_pairwise(epic, "date_infection"),
##   dist = "gamma",
##   max_value = 20,
##   bootstraps = 1
## )

## ## export for caching
## export(
##   generation_time,
##   here("data/cache/epidemic_models/generation_time.rds")
## )

## import cached generation time
generation_time <- import(here("data/cache/epidemic_models/generation_time.rds"))

## ## estimate incubation period
## incubation_period <- bootstrapped_dist_fit(
##   linelist$date_onset - linelist$date_infection,
##   dist = "lognormal",
##   max_value = 100,
##   bootstraps = 1
## )

## ## export for caching
## export(
##   incubation_period,
##   here("data/cache/epidemic_models/incubation_period.rds")
## )

## import cached incubation period
incubation_period <- import(here("data/cache/epidemic_models/incubation_period.rds"))

## get incidence from onset date
cases <- linelist %>%
  group_by(date = date_onset) %>%
  summarise(confirm = n())

## ## run epinow
## epinow_res <- epinow(
##   reported_cases = cases,
##   generation_time = generation_time,
##   delays = delay_opts(incubation_period),
##   target_folder = here("data/cache/epidemic_models"),
##   return_output = TRUE,
##   output = "samples",
##   verbose = TRUE,
##   stan = stan_opts(samples = 750, chains = 4),
##   horizon = 21
## )

## ## export for caching
## export(
##   epinow_res,
##   here("data/cache/epidemic_models/epinow_res.rds")
## )

## import cached epinow results
epinow_res <- import(here("data/cache/epidemic_models/epinow_res.rds"))

## plot summary figure
plot(epinow_res)

```

<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}

We will use two different methods and packages for R<sub>t</sub> estimation,
namely **EpiNow** and **EpiEstim**, as well as the **projections** package for
forecasting case incidence.
	
```{r epidemic_models_packages, eval = TRUE}
pacman::p_load(
   rio,          # File import
   here,         # File locator
   tidyverse,    # Data management + ggplot2 graphics
   epicontacts,  # Analysing transmission networks
   EpiNow2,      # Rt estimation
   EpiEstim,     # Rt estimation
   projections   # Incidence projections
)
```
	
We will use the standard, cleaned linelist for all analyses in this section.

```{r transmission_chains_load_show, eval=F}
# import the cleaned linelist
linelist <- rio::import("linelist_cleaned.xlsx")
```


<!-- ======================================================= -->
## Estimating R<sub>t</sub> {.tabset .tabset-fade .tabset-pills}

### EpiNow2 vs. EpiEstim

The reproduction number R is a measure of the transmissibility of a disease and
is defined as the expected number of secondary cases per infected case. In a
fully susceptible population, this value represents the basic reproduction
number R<sub>0</sub>. However, as the number of susceptible individuals in a
population changes over the course of an outbreak or pandemic, and as various
response measures are implemented, the most commonly used measure of
transmissibility is the effective reproduction number R<sub>t</sub>; this is
defined as the expected number of secondary cases per infected case at a given
time _t_.

The **EpiNow2** package provides the most sophisticated framework for estimating
R<sub>t</sub>. It has two key advantages over the other commonly used package,
**EpiEstim**:

* It accounts for delays in reporting and can therefore estimate R<sub>t</sub>
  even when recent data is incomplete.
* It estimates R<sub>t</sub> on _dates of infection_ rather than the dates of
  onset of reporting, which means that the effect of an intervention will
  be immediately reflected in a change in R<sub>t</sub>, rather than with a
  delay.

However, it also has two key disadvantages:

* It requires knowledge of the generation time distribution (i.e. distribution
  of delays between infection of a primary and secondary cases), incubation
  period distribution (i.e. distribution of delays between infection and symptom
  onset) and any further delay distribution relevant to your data (e.g. if you
  have dates of reporting, you require the distribution of delays from symptom
  onset to reporting). While this will allow more accurate estimation of
  R<sub>t</sub>, **EpiEstim** only requires the serial interval distribution
  (i.e. the distribution of delays between symptom onset of a primary and a
  secondary case), which may be the only distribution available to you.
* **EpiNow2** is significantly slower than **EpiEstim**, anecdotally by a factor
  of about 100. For example, estimating R<sub>t</sub> for the sample outbreak
  considered in this section takes about 4 hours. This may be unfeasible if you
  are regularly updating your R<sub>t</sub> estimates.
  
Which package you choose to use will therefore depend on the data, time and
computational resources available to you.

### EpiNow2

#### Estimating delay distributions

The delay distributions required to run **EpiNow2** depend on the data you
have. Essentially, you need to be able to describe the delay from the date of
infection to the date of the event you want to use to estimate R<sub>t</sub>. If
you are using dates of onset, this would simply be the incubation period
distribution. If you are using dates of reporting, you require the
delay from infection to reporting. As this distribution is unlikely to be known
directly, **EpiNow2** lets you chain multiple delay distributions together; in
this case, the delay from infection to symptom onset (e.g. the incubation
period, which is likely known) and from symptom onset to reporting (which you
can often estimate from the data).

As we have the dates of onset for all our cases in the example linelist, we will
only require the incubation period distribution to link our data (e.g. dates of
symptom onset) to the date of infection. We can either estimate this distribution
from the data or use values from the literature.

A literature estimate of the incubation period of Ebola (taken
from [this paper](https://www.nejm.org/doi/full/10.1056/nejmoa1411100)) with a
mean of 9.1, standard deviation of 7.3 and maximum value of 30 would be
specified as follows:

```{r epidemic_models_incubation_literature, eval=F}
incubation_period_lit <- list(
  mean = log(9.1),
  mean_sd = log(0.1),
  sd = log(7.3),
  sd_sd = log(0.1),
  max = 30
)
```
Note that **EpiNow2** requires these delay distributions to be provided on a **log**
scale, hence `log` call around each value (except the `max` parameter which,
confusingly, has to be provided on a natural scale). The `mean_sd` and `sd_sd`
define the standard deviation of the mean and standard deviation estimates. As
these are not known in this case, we choose the fairly arbitrary value of 0.1.

For our analysis, we will instead estimate the incubation period distribution
from the linelist itself using the function `bootstrapped_dist_fit`, which will
fit a lognormal distribution to the observed delays between infection and onset
in the linelist.

```{r epidemic_models_incubation_estimate, eval=F}
## estimate incubation period
incubation_period <- bootstrapped_dist_fit(
  linelist$date_onset - linelist$date_infection,
  dist = "lognormal",
  max_value = 100,
  bootstraps = 1
)
```

The other distribution we require is the generation time. As we have estimates
of both infection times and transmission links, we can also estimate this
distribution from the linelist by calculating the delay between infection times
of infector-infectee pairs. To do this, we use the handy `get_pairwise` function
from the package **epicontacts**, which allows us to calculate pairwise
differences of linelist properties between transmission pairs. We first create an
epicontacts object (see Transmission chains chapter for further
details):

```{r epidemic_models_epicontacts, eval=F}
## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)
```

We then fit the difference in infection times between transmission pairs,
calculated using `get_pairwise`, to a gamma distribution:

```{r epidemic_models_generation_estimate, eval=F}
## estimate gamma generation time
generation_time <- bootstrapped_dist_fit(
  get_pairwise(epic, "date_infection"),
  dist = "gamma",
  max_value = 20,
  bootstraps = 1
)
```

#### Running **EpiNow2**

Now, we just need to calculate daily incidence values from the linelist. Note
that the column names must be `date` and `confirm`.

```{r epidemic_models_cases, eval=F}
## get incidence from onset date
cases <- linelist %>%
  group_by(date = date_onset) %>%
  summarise(confirm = n())
```

We can then estimate R<sub>t</sub> using the `epinow` function. Some notes on
the inputs:

* We can provide any number of 'chained' delay distributions to the `delays`
  argument; we would simply insert them alongside the `incubation_period` object
  within the `delay_opts` function.
* `return_output` ensures the output is returned within R and not just saved to
  a file.
* `verbose` specifies that we want a readout of the progress.
* `horizon` indicates how many days we want to project future incidence for.
* We pass additional arguments to `stan` specifying how how long we want to run
  the inference for. Increasing `samples` and `chains` will give you a more
  accurate estimate that better characterises uncertainty, however will take
  longer to run.

```{r epidemic_models_run_epinow, eval=F}
## run epinow
epinow_res <- epinow(
  reported_cases = cases,
  generation_time = generation_time,
  delays = delay_opts(incubation_period),
  return_output = TRUE,
  verbose = TRUE,
  horizon = 21,
  stan = stan_opts(samples = 750, chains = 4)
)
```

#### Analysing outputs

Once the results are run, we can plot a summary very easily as follows:

```{r epidemic_models_plot_epinow, eval=T}
## plot summary figure
plot(epinow_res)
```

### EpiEstim

<!-- ======================================================= -->
## Projecting incidence


<!-- ======================================================= -->
## Resources {.tabset .tabset-fade .tabset-pills}

This tab should stay with the name "Resources".
Links to other online tutorials or resources.



