
# Grouping/aggregating data {.tabset .tabset-fade}  
     
This page reviews how to group and aggregate data for descriptive analysis. It makes use of tidyverse packages for common and easy-to-use functions. 

<!-- ======================================================= -->
## Overview {.tabset .tabset-fade .tabset-pills}

Before doing descriptive analyses, it will almost always be a necessary to first group your data and summarize it across these groups (whether it be by time period, place, or a relevant categorical variable) since most often summary statistics across these groups are more meaningful. Luckily, tidyverse makes this really easy through the `group_by` function.

This page will how to perform these grouping operations

* Fast & easy using `group_by()` command in **dplyr** or
* Base R `aggregate()` command 


.drop=F in `group_by()` command


<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}
     
For this tab we use the `linelist` dataset that is cleaned in the Cleaning tab.

**Load packages**  
     
Ensure tidyverse is installed, which includes dplyr for group_by 

```{r}
pacman::p_load(rio,       # to import data
               here,      # to locate files
               tidyverse  # to clean, handle, and plot the data (includes dplyr!)
)
```

Load the data  

```{r, echo=F}
linelist <- rio::import(here("data", "linelist_cleaned.rds"))
```

```{r, eval=F}
linelist <- rio::import(here("data", "linelist_cleaned.xlsx"))
```


<!-- ======================================================= -->
## `group_by()` {.tabset .tabset-fade .tabset-pills}
     
You can perform different operations after first grouping by one variable, say, `outcome`. This provides instruction that any calculations should then be performed within the context of the grouped columns. You can group by 1 or more columns.

First, let's convert `outcome` to a factor to make resulting plots easier to work with.

```{r, eval=F}
linelist <- linelist %>%
  mutate(outcome = as.factor(outcome))
  
```

Below we will walk through a few examples of `group_by` functionalities:

`tally()` gives you a simple count of rows across each category.

```{r, eval=T}
count_by_outcome <- linelist %>%
  group_by(outcome) %>%
  tally()
  
```

Here we see that there were 2 633 deaths, 2 026 recoveries, and 1 348 with no outcome recorded.

We can easily produce summary tables with a range of different descirptive statistics. The `summarise()` after `group_by` allows you to more carefully specify the summary statistic operation to be performed. Below we will find the average age across each outcome group. 

Remember to use na.rm = TRUE to exclude the NA values from the calculation of mean age.

```{r, eval=T}
avg_age_by_outcome <- linelist %>%
  group_by(outcome) %>%
  summarise(avg_age = mean(age, na.rm=TRUE ))
  
```

We see that the average age is roughly stable across outcomes, with those recovering being slightly lower at 14.7 years.


We can also group by more than 1 variable. You can either specify these variables, or use the `group_by_at` or `group_by_if` to use specified criteria in which to choose the grouping parameteres.

For instance, we can find the number of cases, by gender and month of onset...

```{r, eval=T}

count_gender_by_month_of_onset <- linelist %>%
  mutate(month_of_onset = format(date_onset,"%B")) %>%
  group_by(month_of_onset, gender) %>%
  tally()
  
```

We could also take initial records from each group, for instance, which can be handy if used in conjunction with sorting. Below we can sort by `date_of_onset` and then find the first case for each `hospital`

```{r, eval=T}
first_record_per_hosp <- linelist %>%
  arrange(date_onset) %>%
  group_by(hospital) %>%
  slice(1)
  
```

You can perform any summary function on grouped data; see the Cheat Sheet here for more info:
https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf

<!-- ======================================================= -->
## `mutate` with grouped data {.tabset .tabset-fade .tabset-pills}

To retain all of the other columns and just add a new variable for average age, we can use `mutate` instead of summarize. This could be perhaps be useful for some additional descriptive statistics where you wanted other variables still intact.

```{r, eval=T}
avg_age_by_outcome_2 <- linelist %>%
  group_by(outcome) %>%
  mutate(avg_age = mean(age, na.rm=TRUE ))
  
```


<!-- ======================================================= -->
## `aggregate()` {.tabset .tabset-fade .tabset-pills}



<!-- ======================================================= -->
## Resources {.tabset .tabset-fade .tabset-pills}

Here are some useful resources for more information:
*https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf
*https://datacarpentry.org/R-genomics/04-dplyr.html
*https://dplyr.tidyverse.org/reference/group_by.html
*https://itsalocke.com/files/DataManipulationinR.pdf
