
# Grouping data { }  
     
This page reviews how to group and aggregate data for descriptive analysis. It makes use of tidyverse packages for common and easy-to-use functions. 




<!-- ======================================================= -->
## Overview {  }

Grouping data is a core component of data management and analysis. Grouped data can be plotted, or summarised by group (whether by time period, place, or a relevant categorical variable). Functions from the **dplyr** package (part of the **tidyverse**) make grouping and subsequent operations quite easy.  

This page will how to perform these grouping operations using

* `group_by()` command in **dplyr**  
  * Grouping  
  * Un-group  
  * Summarise with statistics  
  * Count and tally  
  * Arranging grouped data  
  * Filters applied to grouped data  
  * Mutate columns within grouped data  
  * Select columns within grouped data  
* Base R `aggregate()` command as an alternative  




<!-- ======================================================= -->
## Preparation {  }
     
**Load packages**  
     
Ensure **tidyverse** package is installed and loaded (includes **dplyr**).  

```{r}
pacman::p_load(rio,       # to import data
               here,      # to locate files
               tidyverse, # to clean, handle, and plot the data (includes dplyr)
               janitor    # adding total rows and columns
               )
```

**Load data**  

For this page we use the cleaned `linelist` dataset

```{r, echo=F}
linelist <- rio::import(here("data", "linelist_cleaned.rds"))
```

```{r, eval=F}
linelist <- rio::import(here("data", "linelist_cleaned.xlsx"))
```


The first 50 rows of `linelist`:  

```{r message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist,50), rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
```



<!-- ======================================================= -->
## Grouping {  }
     
The function `group_by()` from **dplyr** groups the rows by the unique values in the specified columns. Each unique value contitutes a group (or unique combination of values, if multiple grouping columns are specified). Subsequent changes to the dataset or calculations can then be performed within the context of each unique group.  

For example, the command below takes the linelist and groups the rows by unique values in column `outcome`, saving the output as a new dataframe `ll_by_outcome`. The column name is placed inside the parentheses of the function `group_by()`.  

```{r}
ll_by_outcome <- linelist %>% 
  group_by(outcome)
```

**Note that there is no perceptible change to the dataset** after `group_by()`, *until* another **dplyr** verb such as `mutate()` or `summarise()` is applied on the "grouped" dataframe.  

You can however "see" the groupings by printing the dataframe. When you print a grouped dataframe, you will see it has been transformed into a `tibble` class object (LINK) which, when printed, displays which grouping columns have been applied and how many groups there are - written just above the header row.  

```{r}
# print to see which groups are active
ll_by_outcome
```


### Unique groups  

**The groups created reflect each unique combination of values in the grouping columns.** To see the groups and the number of rows in each group, pass the grouped data to `tally()`.  

See below that there are **three** unique values in the grouping column `outcome`: "Death", "Recover", and `NA`. See that there were `r nrow(linelist %>% filter(outcome == "Death"))` deaths, `r nrow(linelist %>% filter(outcome == "Recover"))` recoveries, and `r nrow(linelist %>% filter(is.na(outcome)))` with no outcome recorded.

```{r}
linelist %>% 
  group_by(outcome) %>% 
  tally()
```

You can group by more than one column. Below, the dataframe is grouped by `outcome` and `gender`, and then tallied. Note how each unique combination of `outcome` and `gender` is registered as its own group - including missing values for either column.   

```{r}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally()
```
### New columns  

You can also create a new grouping column *within* the group_by() statement. This is equivalent to calling `mutate()` before the `group_by()`. For a quick tabulation this style can be handy, but for more clarity in your code consider creating this column in it's own `mutate()` step and then piping to `group_by()`.

```{r}
# group dat based on a binary column created *within* the group_by() command
linelist %>% 
  group_by(
    age_class = ifelse(age >= 18, "adult", "child")) %>% 
  tally(sort = T)
```

### Replace/add grouping columns  

By default if you run `group_by()` on data that are already grouped, the old groups will be removed and the new one(s) will apply. If you want to add new groups to the existing ones, add the argument `.add=TRUE`.  

````{r, eval=F}
# Grouped by outcome
by_outcome <- linelist %>% 
  group_by(outcome)

# Add grouping by gender in addition
by_outcome_gender <- by_outcome %>% 
  group_by(gender, .add = TRUE)
```


## Un-group  

Data that have been grouped will remain grouped until specifically ungrouped via `ungroup()`. If you forget to ungroup, it can lead to incorrect calculations! Below is an example of removing all grouping columns:  

```{r, eval=F}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally() %>% 
  ungroup()
```

You can also remove grouping by only specific columns, by placing the column name inside.  

```{r, eval=F}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally() %>% 
  ungroup(gender)
```


<span style="color: black;">**_NOTE:_** The verb `count()` automatically ungroups the data after counting.</span>



## Summarise  

By applying the **dplyr** verb `summarise()` to grouped data, you can produce summary tables containing descriptive statistics *for each group*.  

Within the summarise statement, provide the name(s) of the new summary column(s), an equals sign, and then a statistical function to apply to the data, as shown below. Within a statistical function, list the column to be operated on and any relevant arguements. For example, do not forget `na.rm=TRUE` to remove missing values from calculations!  

Below is an example of `summarise()` *applied without grouped data*. The statistics returned are produced from the entire dataset.     

```{r}
linelist %>% 
  summarise(
    mean_age = mean(age_years, na.rm=T),
    max_age  = max(age_years, na.rm=T),
    min_age  = min(age_years, na.rm=T))
```

In contrast, below is the same `summarise()` statement applied to grouped data. The statistics are calculated for each `outcome` group.  

```{r}
linelist %>% 
  group_by(outcome) %>% 
  summarise(
    mean_age = mean(age_years, na.rm=T),
    max_age  = max(age_years, na.rm=T),
    min_age  = min(age_years, na.rm=T))
```

<span style="color: darkgreen;">**_TIP:_** Summarise works with both UK and US spelling - `summarise()` and `summarize()` call the same function.</span>



### Summarise `across()` multiple columns  

You can use summarise across multiple columns using `across()`. Provide a vector of column names, or use the same semantic helper functions used in `select()` (look below) to specify columns by or by criteria.      

Below, `mean()` is applied to ungrouped data (global calculation). The columns are specified, a function is specified (no parentheses), and finally, any additional arguments for the function (e.g. `na.rm=TRUE`).  

```{r}
linelist %>% 
  summarise(across(.cols = c(age_years, temp),
                   .fns = mean,
                   na.rm=T))
```
Below, the same summarise across call is applied on grouped data:  

```{r}
linelist %>% 
  group_by(outcome) %>% 
  summarise(across(.cols = c(age_years, temp), .fns = mean, na.rm=T))
```

Here are those `select()` helper functions that you can place *within* `across()`:  


There are helpers available to assist you in specifying columns:  

* `everything()`  - all other columns not mentioned  
* `last_col()`    - the last column  
* `where()`       - applies a function to all columns and selects those which are TRUE  
* `starts_with()` - matches to a specified prefix. Example: `select(starts_with("date"))`
* `ends_with()`   - matches to a specified suffix. Example: `select(ends_with("_end"))`  
* `contains()`    - columns containing a character string. Example: `select(contains("time"))` 
* `matches()`     - to apply a regular expression (regex). Example: `select(contains("[pt]al"))`  
* `num_range()`   - 
* `any_of()`      - matches if column is named. Useful if the name might not exist. Example: `select(any_of(date_onset, date_death, cardiac_arrest))`  


For example, to return the mean of every numeric column:  

```{r}
linelist %>% 
  group_by(outcome) %>% 
  summarise(across(where(is.numeric), .fns = mean, na.rm=T))
```


If you want summary multiple statistics of multiple columns, in an easy-to-read format, consider a two-way table with the **gtsummary** package This package is demonstrated more extensively in the the Statistics page (LINK).  

```{r, message = F}
library(gtsummary) 

linelist %>% 
  select(outcome, age_years, temp, ht_cm) %>%     # select columns (optional)
  gtsummary::tbl_summary( 
    by = outcome,                                 # indicate grouping column (optional)
    statistic = all_continuous() ~ "{mean} ({sd})") # return mean and std deviation for each group
    
```





## Count and tally  

`count()` and `tally()` provide similar functionality but are different.  

`tally()` is shorthand for `summarise()`, and *does not* automatically group data. Thus, to achieve grouped tallys it must follow a `group_by()` command. You can add `sort = TRUE` to see the largest groups first.    

```{r}
linelist %>% 
  tally
```
```{r}
linelist %>% 
  group_by(outcome) %>% 
  tally(sort = TRUE)
```

In contrast, `count()` does the following:  

* applies `group_by()` on the specified column(s)  
* applies `summarise()` and returned column `n` with the number of observations per group  
* applies `ungroup()`  

```{r}
linelist %>% 
  count(outcome)
```

Just like with `group_by()` you can create a new column within the `count()` command:  

```{r}
linelist %>% 
  count(age_class = ifelse(age >= 18, "adult", "child"), sort = T)
```
Read more about the distinction between `tally()` and `count()` [here](https://dplyr.tidyverse.org/reference/tally.html)  

Both of these verbs can be called multiple times, with the functionality "rolling up". For example, to summarise the number of genders present for each outcome, run the following. Note, the name of the final column is changed from default "n" for clarity.  

```{r}
linelist %>% 
  # produce counts by outcome-gender groups
  count(outcome, gender) %>% 
  # produce counts of gender within each outcome group
  count(outcome, name = "number of genders per outcome" ) 
```


### Add totals  

If you want to add total rows or column after using `tally()` or `count()`, consider using the **janitor** package, which offers functions like `adorn_totals()` and `adorn_percentages()`. There are many useful functions (search Help for details), here are a few of them:  

* Use `adorn_totals()` to get totals - specify the argument `where = ` either "row" or "col" or `c("row", "col")`.  
* Use `adorn_percentages()` to convert counts to proportions - specify the argument `denominator = ` either "row", "col", or "all".  
* Use `adorn_pct_formatting()` to convert proportions to percentages (can specify number of `digits = `, whether to add "%" with `affix_sign =`, and specify specific column names to operate on)  
* Use `adorn_ns()` to add back the underlying counts ("N"s) to a table whose proportions were calculated by `adorn_percentages()` - to display them together. Indicate `position = ` of the Ns as either "rear" or "front" of the proportions.  


To add totals:  

```{r}
linelist %>% 
  count(outcome) %>% 
  adorn_totals(where = "col")
```

To convert the numbers to proportions:  

```{r}
linelist %>% 
  count(outcome) %>% 
  adorn_totals(where = "row") %>%              # add total row
  adorn_percentages(denominator = "col") %>%   # convert to proportions
  adorn_rounding(digits = 2)                   # round the proportions
```

**janitor** functions can be use together, as below:  

```{r}
linelist %>% 
  count(outcome) %>%              # produce the counts by unique outcome
  adorn_totals(where = "row") %>% # add total row
  adorn_percentages("col") %>%    # add proportion by column
  adorn_pct_formatting() %>%      # proportion converted to percent
  adorn_ns(position = "front")    # Add the underlying N, in front of the percentage
```





## Arranging grouped data

TO DO

Using the **dplyr** verb `arrange()` to order the rows in a dataframe behaves the same when the data are grouped, *unless you set the argument `.by_group =TRUE`. In this case the rows are ordered first by the grouping columns and then by any other columns you specify.   






## Filter on grouped data

### `filter()`  

When applied in conjunction with functions that evaluate the dataframe (like `max()`, `min()`, `mean()`), these functions will now be applied to the groups. For example, if you want to filter and keep rows where patients are above the median age, this will now apply per group. 

TO DO MORE


### Slice rows per group  

The **dplyr** function `slice()`, which [subsets rows based on their position](https://dplyr.tidyverse.org/reference/slice.html) in the data, can also be applied per group. Remember to account for sorting the data within each group to get the desired "slice".  

For example, to retrieve only the latest 5 admissions from each hospital:  

1) Group the linelist by column `hospital`  
2) Arrange the records from latest to earliest `date_hospitalisation` *within each hospital group*  
3) Slice to retrieve the first 5 rows  

```{r, eval=T}
linelist %>%
  group_by(hospital) %>%
  arrange(hospital, date_hospitalisation) %>%
  slice_head(n = 5) %>% 
  arrange(hospital) %>% 
  select(case_id, hospital, date_hospitalisation)
```
`slice_head()` - selects n rows from the top  
`slice_tail()` - selects n rows from the end  
`slice_sample()` - randomly selects n rows  
`slice_min()` - selects n rows with highest values in `order_by = ` column, use `with_ties = TRUE` to keep ties  
`slice_max()` - selects n rows with lowest values in `order_by = ` column, use `with_ties = TRUE` to keep ties  




### Filter on group size  

The function `add_count()` adds a column `n` to the original data giving the number of rows in that row's group. 

Shown below for simplicity is a selection of the `linelist` data - `add_count()` is applied to `hospital`, so the values in column `n` reflect the number of rows in that row's hospital group. Note how values are repeated. In the example below, the column name `n` could be changed to `name = `     

```{r}
linelist %>% 
  select(case_id, hospital) %>% 
  add_count(hospital) %>%          # add "number of rows admitted to same hospital as this row" 
  head(10)                          # show just the first 10 rows, for demo purposes
```

It then becomes easy to filter for case rows who were hospitalized at a "small" hospital, say, a hospital that admitted fewer than 500 patients:  

```{r, eval=F}
linelist %>% 
  select(case_id, hospital) %>% 
  add_count(hospital) %>% 
  filter(n < 500)
```





## Mutate on grouped data  

To retain all columns and rows (not summarize) and *add a new variable for average group statistics*, use `mutate()` instead of `summarise()`. 

This is useful if you want group statistics in the original dataset with all other column present - e.g. for calculations comparing one row to the group.  

For example, this code below calculates the difference between a row's delay-to-admission and the median delay for their hospital. The steps are:  

1) Group the data by hospital  
2) Use the column `days_onset_hosp` (delay to hospitalisation) to create a new column containing the mean delay at the hospital of *that row*  
3) Calculate the difference between the two columns  


```{r}
linelist %>% 
  # group data by hospital (no change to linelist yet)
  group_by(hospital) %>% 
  
  # new columns
  mutate(
    # mean days to admission per hospital (rounded to 1 decimal)
    group_delay_admit = round(mean(days_onset_hosp, na.rm=T), 1),
    
    # difference between row's delay and mean delay at their hospital (rounded to 1 decimal)
    diff_to_group     = round(days_onset_hosp - group_delay_admit, 1)) %>%
  
  # select certain rows only - for demonstration/viewing purposes
  select(case_id, hospital, days_onset_hosp, group_delay_admit, diff_to_group)
```



## Select on grouped data  

The verb `select()` works on grouped data, but the grouping columns are always included (even if not mentioned in `select()`).  
If you do not want these grouping columns, use `ungroup()` first.  









<!-- ======================================================= -->
## Base R {  }

Here we briefly demonstrate grouping data with the **base** R function `aggregate()`

TO DO




<!-- ======================================================= -->
## Resources {  }

Here are some useful resources for more information:
*https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf
*https://datacarpentry.org/R-genomics/04-dplyr.html
*https://dplyr.tidyverse.org/reference/group_by.html
https://dplyr.tidyverse.org/articles/grouping.html  
*https://itsalocke.com/files/DataManipulationinR.pdf


You can perform any summary function on grouped data; see the Cheat Sheet here for more info:
https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf



