# Import and export {}

In this page we describe ways to locate, import, and export files:  

* Using the **rio** package to `import()` and `export()` data  
* The **here** package to locate files in your computer or R project  
* Specific import scenarios, such as  
  * Excel sheets  
  * Google sheets 
  * Websites  
  * Skipping rows  
* Exporting/saving data files  


<!-- ======================================================= -->
## Overview

When you import a "dataset" into R, you are generally creating a new data frame object in your R environment and defining it as the imported flat file (Excel, CSV, etc.). To learn more about objects and the assignment operator, see the page on [R Basics].  


<!-- ======================================================= -->
## The **rio** package {}  

The R package we recommend is: **rio**. The `import()` function from **rio** utilizes the file extension (e.g. .xlsx, .csv, .rds, .tsv) to correctly import or export the file. The name "rio" is an abbreviation of "R I/O" (input/output).  

The alternative to using **rio** is to use functions from many other packages, each of which is specific to a type of file. For example, `read.csv()` (**base** R), `read.xlsx()` (**openxlsx** package), and `write_csv()` (**readr** pacakge), etc. These alternatives can be difficult to remember, whereas using `import()` and `export()` from **rio** is easy.  

**rio**'s functions `import()` and `export()` use the appropriate package and function for a given file, based on its file extension. See the end of this page for a complete table of which packages/functions **rio** uses in the background. It can also be used to import STATA, SAS, and SPSS files, among dozens of others.  

Import/export of shapefiles requires other packages, as detailed in the page on [GIS basics].    


<!-- ======================================================= -->
## File paths  

When importing or exporting data, you must provide a file path. You can do this one of three ways:  

1) Provide the "full" / "absolute" file path  
2) Provide a "relative" file path - the location *relative to* an R project root directory  
3) Manual file selection  



### "Absolute" file paths {-}  

This is an example of an absolute file path, placed within quotes and provided to `import()`, to be saved in R as the data frame object named `linelist`.  

```{r, eval=F}
linelist <- import("C:/Users/Pierre/My Documents/epiproject/data/linelists/ebola_linelist.xlsx")
```

A few things to note about absolute file paths:  

* **Avoid using absolute file paths** as they will usually break if the script is run on a different computer
* They can be used if you must load data from a *shared drive* folder distant from where your R script is saved  
* Use *forward* slashes as in the example above (this is *NOT* the default for Windows file paths)  
* File paths that begin with double slashes (e.g. "//...") will likely **not be recognized by R** and will produce an error. Consider moving to a "named" or "lettered" drive that begins with a letter (e.g. "J:" or "C:"). See the page on [Directory interactions] for more details on this issue.  



### "Relative" file paths {-}

Below, the same file location is given *relative to* an R project root folder "epiproject". This assumes you are working in an R project and have loaded the package **here**.  

```{r, eval=F}
linelist <- import(here("data", "linelists", "ebola_linelist.xlsx"))
```

The package **here** and its function `here()` locate files on your computer in relation to the root directory of an R project, and provide a shortcut syntax for providing a file path to a function like `import()`. This is how `here()` works:  

* Ensure you are working within an R project (read more on the [R projects] page  
* When the **here** package is first loaded within the R project, it places a small file called "here" in the root-level folder of your R project as a "benchmark" or "anchor" for all other files in the project.  
* In your script, if you want to reference a file saved in your project’s folders, you use the function `here()` to tell R where the file is located *in relation to that anchor*.  
* `here()` can be used for both importing and exporting.  

If you are unsure where “here” root is set to, run the function `here()` with empty parentheses:

```{r, eval=T}
# load the package
pacman::p_load(here)

# return the folder path that "here" is set to 
here()
```

You can build onto that anchor by specifying further folders, within quotes, separated by commas, finally ending with the filename and extension. This approach also removes slash direction as a source of error. 

Running the `here()` command with folder names and file name returns the extended file path, which can then processed by the `import()` function.

```{r}
# the filepath
here("data", "linelist.xlsx")
```



<!-- ======================================================= -->
### Select file manually {-}

You can import data manually via one of these methods:  

1) Environment RStudio Pane, click "Import Dataset", and select the type of data 
2) Click File / Import Dataset / (select the type of data)  
3) To hard-code manual selection, use the *base R* command `file.choose()` (leaving the parentheses empty) to trigger appearance of a **pop-up window** that allows the user to manually select the file from their computer. For example:  

```{r import_choose, eval=F}
# Manual selection of a file. When this command is run, a POP-UP window will appear. 
# The file path selected will be supplied to the import() command.

my_data <- import(file.choose())
```

<span style="color: darkgreen;">**_TIP:_** The **pop-up window** may appear BEHIND your RStudio window.</span>





<!-- ======================================================= -->
## Excel sheets {}

If you want to import a specific **sheet** from an Excel workbook, include the sheet name to the `which = ` argument. For example:  

```{r eval=F}
my_data <- import("my_excel_file.xlsx", which = "Sheetname")
```

If using the `here()` method to provide a relative pathway to `import()`, you can still indicate a specific sheet by adding the `which = ` argument after the closing parentheses of the `here()` function.  

```{r import_sheet_here, eval=F}
# Demonstration: importing a specific Excel sheet when using relative pathways with the 'here' package
linelist_raw <- import(here("data", "linelist.xlsx"), which = "Sheet1")`  
```

To *export* a dataframe from R to a specific Excel sheet and have the rest of the Excel workbook remain unchanged, you will have to import, edit, and export with an alternative package catered to this purpose such as **openxlsx**. See more information in the page on [Directory interactions] or [at this github page](https://ycphs.github.io/openxlsx/).

If your Excel workbook is .xlsb (binary format Excel workbook) you may not be able to import it using **rio**. Consider re-saving it as .xlsx, or using a package like **readxlsb** which is built for [this purpose](https://cran.r-project.org/web/packages/readxlsb/vignettes/read-xlsb-workbook.html).  






<!-- ======================================================= -->
## Missing values  

You may want to designate which value(s) in your dataset should be considered as missing. As explained in the page on [Missing data], the value in R for missing data is `NA`, but perhaps the dataset you want to import uses 99, "Missing", or just empty character space "" instead.  

Use the `na = ` argument for `import()` and provide the value(s) within quotes (even if they are numbers). You can specify multiple values by including them within a vector, using `c()` as shown below.  

```{r, eval=F}
linelist <- import(here("data", "linelist_raw.xlsx"), na ="99")
```

```{r, eval=F}
linelist <- import(here("data", "cleaning_dict.csv"), na = c("Missing", "", " "))
```



<!-- ======================================================= -->
## Google sheets {}

You can import data from an online Google spreadsheet with the **googlesheet4** package and by authenticating your access to the spreadsheet.  


```{r, eval=F}
pacman::p_load("googlesheets4")
```

Below, a demo Google sheet is imported and saved. This command may prompt confirmation of authentification of your Google account. Follow prompts and pop-ups in your internet browser to grant Tidyverse API packages permissions to edit, create, and delete your spreadsheets in Google Drive.  


The sheet below is "viewable for anyone with the link" and you can try to import it.  

```{r, eval=F}
Gsheets_demo <- read_sheet("https://docs.google.com/spreadsheets/d/1scgtzkVLLHAe5a6_eFQEwkZcc14yFUx1KgOMZ4AKUfY/edit#gid=0")
```

The sheet can also be imported using only the sheet ID, a shorter part of the URL:  

```{r, eval=F}
Gsheets_demo <- read_sheet("1scgtzkVLLHAe5a6_eFQEwkZcc14yFUx1KgOMZ4AKUfY")
```


Another package, **googledrive** offers useful functions for writing, editing, and deleting Google sheets. For example, using the  `gs4_create()` and `sheet_write()` functions found in this package. 

Here are some other helpful online tutorials: [basic importing tutorial](https://arbor-analytics.com/post/getting-your-data-into-r-from-google-sheets/) [more detail](https://googlesheets4.tidyverse.org/articles/googlesheets4.html) [interaction between the two packages](https://googlesheets4.tidyverse.org/articles/articles/drive-and-sheets.html)


<!-- ======================================================= -->
## Scraping websites {}

Scraping data from a website - TBD - Under construction




<!-- ======================================================= -->
## Skip rows  

Sometimes, you may want to avoid importing a row of data. You can do this with the argument `skip = ` if using `import()` from **rio** on a .xlsx or .csv file. Provide the number of rows you want to skip. 


```{r, eval=F}
linelist_raw <- import("linelist_raw.xlsx", skip = 1)  # does not import header row
```

Unfortunately `skip = ` only accepts one integer value, *not* a range (e.g. "2:10" does not work). To skip import of specific rows that are not consecutive from the top, consider importing multiple times and using `bind_rows()` from **dplyr**. See the example below of skipping only row 2.  



### Remove second header row {-}  

Sometimes, your data may have a *second* row that you want to remove, for example if it is a "data dictionary" row as shown below. This situation can be problematic because it can result in all columns being imported as class "character".  

```{r, echo=F}
# HIDDEN FROM READER
####################
# Create second header row of "data dictionary" and insert into row 2. Save as new dataframe.
linelist_2headers <- rio::import(here::here("data", "linelist_cleaned.rds")) %>%         
        mutate(across(everything(), as.character)) %>% 
        add_row(.before = 1,
                #row_num = "000",
                case_id = "case identification number assigned by MOH",
                generation = "transmission chain generation number",
                date_infection = "estimated date of infection, mm/dd/yyyy",
                date_onset = "date of symptom onset, YYYY-MM-DD",
                date_hospitalisation = "date of initial hospitalization, mm/dd/yyyy",
                date_outcome = "date of outcome status determination",
                outcome = "either 'Death' or 'Recovered' or 'Unknown'",
                gender = "either 'm' or 'f' or 'unknown'",
                hospital = "Name of hospital of first admission",
                lon = "longitude of residence, approx",
                lat = "latitude of residence, approx",
                infector = "case_id of infector",
                source = "context of known transmission event",
                age = "age number",
                age_unit = "age unit, either 'years' or 'months' or 'days'",
                fever = "presence of fever on admission, either 'yes' or 'no'",
                chills = "presence of chills on admission, either 'yes' or 'no'",
                cough = "presence of cough on admission, either 'yes' or 'no'",
                aches = "presence of aches on admission, either 'yes' or 'no'",
                vomit = "presence of vomiting on admission, either 'yes' or 'no'",
                time_admission = "time of hospital admission HH:MM")
```


```{r message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist_2headers, 5), rownames = FALSE, filter="top", options = list(pageLength = 4, scrollX=T), class = 'white-space: nowrap' )
```

To solve this, you will likely need to import the data twice.  

1) Import the data in order to store the correct column names  
2) Import the data again, skipping the first *two* rows (header and second rows)  
3) Bind the correct names onto the reduced dataframe

The exact argument used to bind the correct column names depends on the type of data file (.csv, .tsv, .xlsx, etc.). This is because **rio** is using a different function for the different file types (see table above).  

**For Excel files:** (`col_names = `)  

```{r, eval=F}
# import first time; store the column names
linelist_raw_names <- import("linelist_raw.xlsx") %>% names()  # save true column names

# import second time; skip row 2, and assign column names to argument col_names =
linelist_raw <- import("linelist_raw.xlsx",
                       skip = 2,
                       col_names = linelist_raw_names
                       ) 
```

**For CSV files:** (`col.names = `)  

```{r, eval=F}
# import first time; sotre column names
linelist_raw_names <- import("linelist_raw.csv") %>% names() # save true column names

# note argument for csv files is 'col.names = '
linelist_raw <- import("linelist_raw.csv",
                       skip = 2,
                       col.names = linelist_raw_names
                       ) 
```

**Backup option** - changing column names as a separate command

```{r, eval=F}
# assign/overwrite headers using the base 'colnames()' function
colnames(linelist_raw) <- linelist_raw_names
```

Bonus! If you do have a second row that is a data dictionary, you can easily create a proper data dictionary from it. See this [vignette](https://alison.rbind.io/post/2018-02-23-read-multiple-header-rows/)  


### Make a data dictionary {-}  

Bonus! If you do have a second row that is a data dictionary, you can easily create a proper data dictionary from it. This tip is adapted from this [post](source: https://alison.rbind.io/post/2018-02-23-read-multiple-header-rows/).  


```{r}
dict <- linelist_2headers %>%             # begin: linelist with dictionary as first row
  head(1) %>%                             # keep only column names and first dictionary row                
  pivot_longer(cols = everything(),       # pivot all columns to long format
               names_to = "Column",       # assign new column names
               values_to = "Description")
```


```{r message=FALSE, echo=F}
DT::datatable(dict, rownames = FALSE, filter="top", options = list(pageLength = 4, scrollX=T), class = 'white-space: nowrap' )
```



### Combine two header rows {-}  

In some cases, you may want to combine two header rows into one. This command will define the column names as the combination (pasting together) of the existing column names with the value underneath in the first row. Replace "df" with the name of your dataset.  

```{r, eval=F}
names(df) <- paste(names(df), df[1, ], sep = "_")
```


## Multiple files - import, export, split, combine  

See the page on [Iteration] for examples of how to import and combine multiple files, or multiple Excel workbook files. This page also has examples on how to split a data frame into parts and export each one separately, or as named sheets in an Excel workbook.  



<!-- ======================================================= -->
## Manual data entry {}

### Entry by rows {-}  

Use the `tribble` function from the **tibble** package from the tidverse ([onlinetibble reference](https://tibble.tidyverse.org/reference/tribble.html)).  
  
Note how column headers start with a *tilde* (`~`).  Also note that each column must contain only one class of data (character, numeric, etc.). You can use tabs, spacing, and new rows to make the data entry more intuitive and readable. Spaces do not matter between values, but each row is represented by a new line of code. For example:  

```{r import_manual_row}
# create the dataset manually by row
manual_entry_rows <- tibble::tribble(
  ~colA, ~colB,
  "a",   1,
  "b",   2,
  "c",   3
  )
```

And now we display the new dataset:  

```{r, echo=F}
# display the new dataset
DT::datatable(manual_entry_rows)
```


### Entry by columns {-}  

Since a data frame consists of vectors (vertical columns), the **base** approach to manual dataframe creation in R expects you to define each column and then bind them together. This can be counter-intuitive in epidemiology, as we usually think about our data in rows (as above). 

```{r import_manual_col}
# define each vector (vertical column) separately, each with its own name
PatientID <- c(235, 452, 778, 111)
Treatment <- c("Yes", "No", "Yes", "Yes")
Death     <- c(1, 0, 1, 0)
```

<span style="color: orange;">**_CAUTION:_** All vectors must be the same length (same number of values).</span>

The vectors can then be bound together using the function `data.frame()`:  

```{r}
# combine the columns into a data frame, by referencing the vector names
manual_entry_cols <- data.frame(PatientID, Treatment, Death)
```

And now we display the new dataset:  

```{r, echo=F}
# display the new dataset
DT::datatable(manual_entry_cols)
```




### Pasting from clipboard {-}  


If you copy data from elsewhere and have it on your clipboard, you can try the following function from **base** R to convert those data into an R data frame:  

```{r, eval=F}
df_from_clipboard <- read.table(
  file = "clipboard",  # specify this as "clipboard"
  sep = "t",           # separator could be tab, or commas, etc.
  header=TRUE)         # if there is a header row
```




<!-- ======================================================= -->
## Export {}  

With **rio**, you can use the `export()` function in a very similar way to `import()`. First give the name of the R object you want to save (e.g. `linelist`) and then in quote the filepath including name and file extension. For example:  

```{r, eval=F}
export(linelist, "my_linelist.xlsx") # will save to working directory
```

You could save the same dataframe as a .csv, and to a folder specified by **here** relative pathway:  

```{r, eval=F}
export(linelist, here("data","clean", "my_linelist.csv")
```



## RDS files {}

Along with .csv, .xlsx, etc, you can also export/save R data frames as .rds files. This is a file format specific to R, and is very useful if you know you will work with the exported data again in R. 

The classes of columns are stored, so you don't have do to cleaning again when it is imported (with an Excel or even a CSV file this can be a headache!).  

For example, if you work in an Epi team and need to send files to a GIS team for mapping, and they use R as well, just send them the .rds file! Then all the column classes are retained and they have less work to do.  

```{r, eval=F}
export(linelist, here("data","clean", "my_linelist.rds")
```



<!-- ======================================================= -->
## Rdata files {}

`.Rdata` files store R objects, and can actually store multiple R objects within one file, for example multiple dataframes, model results, lists, etc. This can be very useful to consolidate or share a lot of your data for a given project.  

In the below example, multiple R objects are stored within the exported file "my_objects.Rdata":  

```{r, eval=F}
rio::export(my_list, my_dataframe, my_vector, "my_objects.Rdata")
```

Note: if you are trying to *import* a list, use `import_list()` from **rio** to import it witht the complete original structure and contents.  

```{r, eval=F}
rio::import_list("my_list.Rdata")
```





<!-- ======================================================= -->
## Saving plots {} 

How to save plots, such as those created by `ggplot()` is discussed in depth in the [ggplot tips] page.  

In brief, run `ggsave("my_plot_filepath_and_name.png")` after printing your plot. You can either provide a saved plot object to the `plot = ` argument, or only specify the destination file path (with file extension) to save the most recently-displayed plot. You can also control the `width = `, `height = `, `units = `, and `dpi = `.  

How to save a network graph, such as a transmission tree, is addressed in the page on [Transmission chains]. 


<!-- ======================================================= -->
## Resources {} 

The [R Data Import/Export Manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html)  
[R 4 Data Science chapter](https://r4ds.had.co.nz/data-import.html#data-import)  
[ggsave](https://ggplot2.tidyverse.org/reference/ggsave.html)  


Below is a table, taken from the **rio** online [vignette](https://cran.r-project.org/web/packages/rio/vignettes/rio.html). For each type of data it shows: the expected file extension, the package **rio** uses to import or export the data, and whether this functionality is included in the default installed version of **rio**.  



Format                     | Typical Extension | Import Package    | Export Package     | Installed by Default
---------------------------|-------------------|-------------------|--------------------|---------------------
Comma-separated data | .csv | data.table `fread()` | data.table |	Yes
Pipe-separated data |	.psv | data.table `fread()` | data.table | Yes
Tab-separated data| .tsv | data.table `fread()` | data.table | Yes
SAS | .sas7bdat | haven | haven | Yes
SPSS | .sav | haven | haven | Yes
Stata | .dta | haven | haven | Yes
SAS | XPORT | .xpt | haven | haven | Yes
SPSS Portable | .por | haven | | Yes
Excel | .xls | readxl | | Yes
Excel | .xlsx | readxl | openxlsx | Yes
R syntax | .R	| base | base | Yes
Saved R objects | .RData, .rda | base | base | Yes
Serialized R objects | .rds | base | base | Yes
Epiinfo | .rec | foreign | | Yes
Minitab | .mtp | foreign | | Yes
Systat | .syd |	foreign | | Yes
“XBASE” | database files | .dbf | foreign | foreign | Yes
Weka Attribute-Relation File Format | .arff | foreign | foreign | Yes
Data Interchange Format | .dif | utils | | Yes
Fortran data | no recognized extension | utils | | Yes
Fixed-width format data | .fwf | utils | utils | Yes
gzip comma-separated data | .csv.gz | utils | utils | Yes
CSVY (CSV + YAML metadata header) | .csvy | csvy | csvy | No
EViews | .wf1 |hexView | | No
Feather R/Python interchange format | .feather | feather | feather | No
Fast Storage | .fst | fst |	fst | No
JSON | .json | jsonlite | jsonlite | No
Matlab | .mat | rmatio | rmatio | No
OpenDocument Spreadsheet | .ods | readODS | readODS | No
HTML Tables | .html | xml2 | xml2 | No
Shallow XML documents | .xml | xml2 | xml2 | No
YAML | .yml | yaml | yaml	| No
Clipboard	default is tsv | |  clipr | clipr | No



