
# R on network drives { }  

 


<!-- ======================================================= -->
## Overview {  }

Using R on network or "company" shared drives can be extremely frustrating. This page contains approaches, common errors, and suggestions on troubleshooting, including for the particularly delicate situations involving Rmarkdown.  

**Using R on Network Drives: Overarching principles**  

1) Must have administrator access on your computer. Setup RStudio specifically to run as administrator.  
2) Use your "\\\" package library as little as possible, save packages to "C:" library when possible.  
3) the **rmarkdown** package must not be in a "\\\" library, as then it can't talk to TinyTex or Pandoc.  



<!-- ======================================================= -->
## Preparation {  }


**Using R on Network Drives: Overarching principles**  

1) Must have administrator access on your computer. Setup RStudio specifically to run as administrator.  
2) Use your "\\\" package library as little as possible, save packages to "C:" library when possible.  
3) the **rmarkdown** package must not be in a "\\\" library, as then it can't talk to TinyTex or Pandoc.  

**Useful commands**

```{r, eval=F}
# Find libraries
.libPaths()                   # Your library paths, listed in order that R installs/searches. 
                              # Note: all libraries will be listed, but to install to some (e.g. C:) you 
                              # may need to be running RStudio as an administrator (it won't appear in the 
                              # install packages library drop-down menu) 

# Switch order of libraries
# this can effect the priority of R finding a package. E.g. you may want your C: library to be listed first
myPaths <- .libPaths() # get the paths
myPaths <- c(myPaths[2], myPaths[1]) # switch them
.libPaths(myPaths) # reassign them

# Find Pandoc
Sys.getenv("RSTUDIO_PANDOC")  # Find where RStudio thinks your Pandoc installation is

# Find a package
# gives first location of package (note order of your libraries)
find.package("rmarkdown", lib.loc = NULL, quiet = FALSE, verbose = getOption("verbose")) 
```



<!-- ======================================================= -->
## Troubleshooting common errors {  }


**"Failed to compile...tex in rmarkdown"**  

check/install tinytex, to C: location

```{r, eval=F}
# check/install tinytex, to C: location
tinytex::install_tinytex()
tinytex:::is_tinytex() # should return TRUE (note three colons)
```


**Internet routines cannot be loaded**  

For example, "Error in tools::startDynamicHelp() : internet routines cannot be loaded"  

* Try selecting 32-bit version from RStudio via Tools/Global Options.  
  * note: if 32-bit version does not appear in menu, make sure not using RStudio v1.2.  
* Or try uninstalling R and re-installing with different bit (32 instead of 64)


**C: library does not appear as an option when I try to install packages manually**

* Must run RStudio as an administrator, then it will appear.  
* To set-up RStudio to always run as administrator (advantageous when using an Rproject where you don't click RStudio icon to open)... right-click the Rstudio icon, open properties, compatibility, and click the checkbox Run as Administrator.  


**Pandoc 1 error**  

If you are getting pandoc error 1 when knitting Rmarkdowns on network drives:  

* this can help (of two library locations, have the one with lettered drive listed first)  
* This worked when knitting on local drive but while on network internet connection  
* See https://ciser.cornell.edu/rmarkdown-knit-to-html-word-pdf/  

```{r, eval=F}
myPaths <- .libPaths() # get the library paths
myPaths <- c(myPaths[2], myPaths[1]) # switch them
.libPaths(myPaths) # reassign them
```


**Pandoc Error 83 (can't find file...rmarkdown...lua...)**  
This means that it was unable to find this file.  

See https://stackoverflow.com/questions/58830927/rmarkdown-unable-to-locate-lua-filter-when-knitting-to-word  

Possibilities:  

1) Rmarkdown package is not installed  
2) Rmarkdown package is not findable  
3) an admin rights issue.  

R is not able to find the 'rmarkdown' package file, so check which library the rmarkdown package lives. 
If it is in a library that in inaccessible (e.g. starts with "\\\") consider manually moving it to C: or other named drive library.  
But be aware that the rmarkdown package has to be able to reach tinytex, so rmarkdown package can't live on a network drive.


**Pandoc Error 61**
For example: "Error: pandoc document conversion failed with error 61"  

"Could not fetch..."  

* Try running RStudio as administrator (right click icon, select run as admin, see above instructions)  
* Also see if the specific package that was unable to be reached can be moved to C: library.

**LaTex error (see below)**

"! Package pdftex.def Error: File `cict_qm2_2020-06-29_files/figure-latex/unnamed-chunk-5-1.png' not found: using draft setting."

"Error: LaTeX failed to compile file_name.tex."  
See https://yihui.org/tinytex/r/#debugging for debugging tips. 
See file_name.log for more info.


**Pandoc Error 127**
This could be a RAM (space) issue. Re-start your R session and try again. 


**Mapping network drives**

How does one open a file "through a mapped network drive"?  

* First, you'll need to know the network location you're trying to access.  
* Next, in the Windows file manager, you will need to right click on "This PC" on the right hand pane, and select "Map a network drive".  
* Go through the dialogue to define the network location from earlier as a lettered drive.  
* Now you have two ways to get to the file you're opening. Using the drive-letter path should work.  

From: https://stackoverflow.com/questions/48161177/r-markdown-openbinaryfile-does-not-exist-no-such-file-or-directory/55616529?noredirect=1#comment97966859_55616529


**ISSUES WITH HAVING A SHARED LIBRARY LOCATION ON NETWORK DRIVE**  

**Error in install.packages()**  

Try removing... /../.../00LOCK (directory)  

* Manually delete the 00LOCK folder directory from your package the library. Try installing again.  
* You can try the command pacman::p_unlock() (you can also put this command in the Rprofile so it runs every time project opens.)  
* Then try installing the package again. It may take several tries.  
* If all else fails, install the package to another library and then manually copy it over.  



<!-- ======================================================= -->
## Resources {  }

This tab should stay with the name "Resources".
Links to other online tutorials or resources.




