# Univariate and multivariable regression { }

<!-- ======================================================= -->

## Overview {  }

This page demonstrates the use of **base** R regression function such as `glm()` and the **gtsummary** package to 
look at associations between variables (e.g. odds ratios, risk ratios and hazard
ratios). It also uses functions like `tidy()` from the **broom** package to clean-up regression outputs.  

1.  Univariate: two-by-two tables 
2.  Stratified: mantel-haenszel estimates 
3.  Multivariable: variable selection, model selection, final table
4.  Forest plots

<span style="color: black;">**_NOTE:_** We use the term *multivariable* to refer to a regression with multiple explanatory variables. In this sense a *multivariate* model would be a regression with several outcomes - see this [editorial](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3518362/) for detail </span> 

<!-- ======================================================= -->

## Preparation {  }


### Load packages {-}

This code chunk shows the loading of packages required for the analyses. In this handbook we emphasize `p_load()` from **pacman**, which installs the package if necessary and loads it for use. You can also load packages with `library()` from **base** R. See the page on [R basics] for more information on R packages.  

```{r}
pacman::p_load(
  rio,          # File import
  here,         # File locator
  tidyverse,    # data management + ggplot2 graphics, 
  stringr,      # manipulate text strings 
  purrr,        # loop over objects in a tidy way
  gtsummary,    # summary statistics and tests 
  broom,        # tidy up results from regressions
  lmtest,       # likelihood-ratio tests
  parameters,   # alternative to tidy up results from regressions
  see          # alternative to visualise forest plots
  )
```

### Import data {-}
We import the dataset of cases from a simulated Ebola epidemic. If you want to download the data to follow step-by-step, see instructions in the [Download book and data] page. The dataset is imported using the `import()` function from the **rio** package. See the page on [Import and export] for various ways to import data.

```{r, echo=F}
# import the linelist into R
linelist <- rio::import(here::here("data", "linelist_cleaned.rds"))
```

```{r, eval=F}
# import the linelist
linelist <- import("linelist_cleaned.xlsx")
```

The first 50 rows of the linelist are displayed below.

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist, 50), rownames = FALSE, options = list(pageLength = 5, scrollX=T) )
```

### Clean data {-}

**Store the explanatory variables**  

```{r}
## define variables of interest 
explanatory_vars <- c("gender", "fever", "chills", "cough", "aches", "vomit")
```


**Convert to 1's and 0's**  

Below we convert the explanatory columns from "yes"/"no" "m"/"f", and "dead"/"alive" to **1 / 0**, to cooperate with the expectations of logistic regression models. To do this efficiently, we define a vector of the column names of our explanatory variables. 

We apply the function `case_when()` to convert specified values to 1's and 0's. This function is applied all the `explanatory_vars`  columns using `across()` (see page on [Grouping data]).  

```{r}
## convert dichotomous variables to 0/1 
linelist <- linelist %>% 
  mutate(
    ## for each of the variables listed and "outcome"
    across(
      all_of(c(explanatory_vars, "outcome")), 
      ## recode male, yes and death to 1; female, no and recover to 0
      ## otherwise set to missing
           ~case_when(
             . %in% c("m", "yes", "Death")   ~ 1,
             . %in% c("f", "no",  "Recover") ~ 0, 
             TRUE                            ~ NA_real_
           ))
  )
```

**Drop rows with missing values**  

To drop rows with missing values, we add the column `age` to the `explanatory_vars` (`age` would have produced an error in the previous `case_when()` operation). Then we pipe the `linelist` to `drop_na()` to remove any rows with missing values in the `outcome` column or any of the `explanatory_vars` columns.  

```{r}
## add in age_category to the explanatory vars 
explanatory_vars <- c(explanatory_vars, "age_cat")

## drop rows with missing information for variables of interest 
linelist <- linelist %>% 
  drop_na(any_of(c("outcome", explanatory_vars)))

```

The number of rows remaining in `linelist` is `r nrow(linelist)`.  


<!-- ======================================================= -->

## Univariate {  }

Just like in the page on [Descriptive tables](https://epirhandbook.com/descriptive-tables.html), your use case will determine which R package you use. We present two options for doing univariate analysis:  

* Use functions available in **base** to quickly print results to the console. Use the **broom** package to tidy up the outputs.  
* Use the **gtsummary** package to model and get publication-ready outputs  



<!-- ======================================================= -->

### **base** R {-}

The function `glm()` from the **stats** package (part of **base** R) is used to fit Generalized Linear Models (GLM).  

`glm()` can be used for univariate and multivariable logistic regression (e.g. to get Odds Ratios). Here are the core parts:  

```{r, eval=F}
# arguments for glm()
glm(formula, family, data, weights, subset, ...)
```

* `formula = ` The model is provided to `glm()` as an equation, with the outcome on the left and explanatory variables on the right of a tilde `~`. In this example we are assessing the association between different age categories and the outcome of death (now coded as 1, see Preparation section).  
* `family = ` This determines the type of model to run. For logistic regression, use `family = "binomial"`, for poisson use `family = "poisson"`. Other examples are in the table below.  
* `data = ` Specify your data frame  


If necessary, you can also specify the link function via the syntax `family = familytype(link = "linkfunction"))`. You can read more in the documentation about other families and optional arguments such as `weights = ` and `subset = ` (`?glm`).  



Family                 | Default link function 
-----------------------|-------------------------------------------  
`"binomial"` | `(link = "logit")`  
`"gaussian"` | `(link = "identity")`  
`"Gamma"` | `(link = "inverse")`  
`"inverse.gaussian"` | `(link = "1/mu^2")`  
`"poisson"` | `(link = "log")`  
`"quasi"` | `(link = "identity", variance = "constant")`  
`"quasibinomial"` | `(link = "logit")`  
`"quasipoisson"` | `(link = "log")`  


When running `glm()` it is most common to save the results as a named R object. Then you can print the results to your console using `summary()` as shown below, or perform other operations on the results (e.g. exponentiate).  

If you need to run a negative binomial regression you can use the **MASS** package; the `glm.nb()` uses the same syntax as `glm()`. 
For a walk-through of different regressions, see the [UCLA stats page](https://stats.idre.ucla.edu/other/dae/). 

#### Univariate `glm()` {-}

Below is a univariate model of `outcome` by age category. We save the model output as `model` and then print it within `summary()` to the console. Note the Estimates provided are the *log odds*. The baseline level is the first factor level of `age_cat` (0-4).  

```{r}
model <- glm(outcome ~ age_cat, family = "binomial", data = linelist)
summary(model)
```

To alter the baseline level of a given variable, ensure the column is class Factor and set the first level (see page on [Factors]. For example below we take `linelist` column `age_cat` and set "20-29" as the baseline before piping the modified dataset into `glm()`.  

```{r}
linelist %>% 
  mutate(age_cat = fct_relevel(age_cat, "20-29", after = 0)) %>% 
  glm(formula = outcome ~ age_cat, family = "binomial") %>% 
  summary()
```

#### Printing results {-}

For most uses, several modifications must be made to the above outputs. The function `tidy()` from the package **broom** is very convenient for making the model results presentable. Here we demonstrate how to combine model outputs with a table of counts.  

1) Get the *exponentiated* log odds ratio estimates and confidence intervals by passing the model to `tidy()` and setting `exponentiate = TRUE` and `conf.int = TRUE`.  

```{r odds_base_single}

model <- glm(outcome ~ age_cat, family = "binomial", data = linelist) %>% 
  # clean up the outputs of the regression (exponentiate and produce CIs)
  tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
  # round all numeric columns
  mutate(across(is.numeric, round, digits = 2))
```

Below is the outputted tibble `model`:  

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(model, rownames = FALSE, options = list(pageLength = nrow(model), scrollX=T), class = 'white-space: nowrap' )
```

2) Combine these model results with a table of counts. Below, we create the counts table by applying **dplyr** functions on the `linelist` (see page on [Grouping data]).  

* Group rows by outcome, and get counts by age category  
* Pivot wider so the column are `age_cat`, `0`, and `1`  
* Remove row for `NA` `age_cat`, if applicable, to align with the model results  

```{r}

counts_table <- linelist %>% 
  ## remove cases with missing outcome or age category
  filter(!is.na(outcome) & !is.na(age_cat)) %>% 
  ## get counts of variable of interest grouped by outcome
  group_by(outcome) %>% 
  ## gets number or rows by unique outcome-age category combinations  
  count(age_cat) %>% 
  ## spread data to wide format (as in cross-tabulation)
  pivot_wider(names_from = outcome, values_from = n) 
```

Here is what this `counts_table` data frame looks like:  

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(counts_table, rownames = FALSE, options = list(pageLength = nrow(counts_table), scrollX=T), class = 'white-space: nowrap' )
```

Now we can bind the `counts_table` and the `model` results together horizontally with `bind_cols()` (**dplyr**). In this code, the `.` represents the piped object `counts_table` and we bind it to `model`. To finish the process, we use `select()` to pick the desired columns and their order, and apply the **base** R `round()` function to all numeric columns specifying 2 decimal places.  

```{r, message=F, warning=F}
combined <- counts_table %>% 
  ## merge with the outputs of the regression 
  bind_cols(., model) %>% 
  ## only keep columns interested in 
  select(term, 2:3, estimate, conf.low, conf.high, p.value) %>% 
  ## round values to 2 decimal places
  mutate(across(is.numeric, round, digits = 2))
```

Here is what the combined data frame looks like:  


```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(combined, rownames = FALSE, options = list(pageLength = nrow(combined), scrollX=T), class = 'white-space: nowrap' )
```


#### Looping multiple univariate models {-}  

Below we present a method using `glm()` and `tidy()` for a more simple approach, see the section on **gtsummary**.  

To run the models on several exposure variables to produce univariate odds ratios (i.e. 
not controlling for each other), you can use the approach below. It uses `str_c()` from **stringr** to create univariate formulas,, runs the `glm()` regression on each formula, passes each `glm()` output to `tidy()` and finally collapses all the model outputs together with `bind_rows()` from **tidyr***. This approach uses `map()` from the package **purrr** to iterate - see the page on [Iteration and loops] for more information on this tool.  

1) Create a vector of column names of the explanatory variables. We already have this as `explanatory_vars` from the data Preparation section of this page.  

2) Use `str_c()` to create multiple string formulas, with `outcome` on the left, and a column name from `explanatory_vars` on the right. The period `.` substitutes for the column name in `explanatory_vars`.  

```{r}
explanatory_vars %>% str_c("outcome ~ ", .)
```

3) Pass these string formulas to `map()` and set `~glm()` as the function to apply to each input. Within `glm()`, set the regression formula as `as.formula(.x)` where `.x` will be replaced by the string formula defined in the step above. `map()` will loop over each of the string formulas, running regressions for each one.  

4) The outputs of this first `map()` are passed to a second `map()` command, which applied `tidy()` to the regression outputs.  

5) Finally the output of the second `map()` (a list of tidied data frames) is condensed with `bind_rows()`, resulting in one data frame with all the univariate results.  


```{r odds_base_multiple}

models <- explanatory_vars %>%       # begin with variables of interest
  str_c("outcome ~ ", .) %>%         # combine each variable into formula ("outcome ~ variable of interest")
  
  # iterate through each univariate formula
  map(                               
    .f = ~glm(                       # pass the formulas one-by-one to glm()
      formula = as.formula(.x),      # within glm(), the string formula is .x
      family = "binomial",           # specify type of glm (logistic)
      data = linelist)) %>%          # dataset
  
  # tidy up each of the glm regression outputs from above
  map(
    .f = ~tidy(
      .x, 
      exponentiate = TRUE,           # exponentiate 
      conf.int = TRUE)) %>%          # return confidence intervals
  
  # collapse the list of regression outputs in to one data frame
  bind_rows() %>% 
  
  # round all numeric columns
  mutate(across(is.numeric, round, digits = 2))
```

This time, the end object `models` is longer because it now represents the combined results of several univariate regressions. Click through to see all the rows of `model`.  

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(models, rownames = FALSE, options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap' )
```

As before, we can create a counts table from the `linelist` for each explanatory variable, bind it to `models`, and make a nice table. We begin with the variables, and iterate through them with `map()`. We iterate through a user-defined function which involves creating a counts table with **dplyr** functions. Then the results are combined and bound with the `models` model results.  


```{r, warning=F, message=F}

## for each explanatory variable
univ_tab_base <- explanatory_vars %>% 
  map(.f = 
    ~{linelist %>%                ## begin with linelist
        group_by(outcome) %>%     ## group data set by outcome
        count(.data[[.x]]) %>%    ## produce counts for variable of interest
        pivot_wider(              ## spread to wide format (as in cross-tabulation)
          names_from = outcome,
          values_from = n) %>% 
        filter(!is.na(.data[[.x]])) %>%  ## drop rows with missings
        rename("variable" = .x) %>%      ## change variable of interest column to "variable"
        mutate(variable = as.character(variable))} ## convert to character, else non-dichotomous (categorical) variables come out as factor and cant be merged
      ) %>% 
  
  ## collapse the list of count outputs in to one data frame
  bind_rows() %>% 
  
  ## merge with the outputs of the regression 
  bind_cols(., models) %>% 
  
  ## only keep columns interested in 
  select(term, 2:3, estimate, conf.low, conf.high, p.value) %>% 
  
  ## round decimal places
  mutate(across(is.numeric, round, digits = 2))

```

Below is what the data frame looks like. See the page on [Tables for presentation] for ideas on how to further convert this table to pretty HTML output (e.g. with **flextable**).  

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(univ_tab_base, rownames = FALSE, options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap' )
```





<!-- ======================================================= -->

### **gtsummary** package {-}

Below we present the use of `tbl_uvregression()` from the **gtsummary** package. Just like in the page on [Descriptive tables](https://epirhandbook.com/descriptive-tables.html), **gtsummary** functions do a good job of running statistics *and* producing professional-looking outputs. This function produces a table of univariate regression results.  

We select only the necessary columns from the `linelist` (explanatory variables and the outcome variable) and  pipe them into `tbl_uvregression()`. We are going to run univariate regression on each of the columns we defined as `explanatory_vars` in the data Preparation section (gender, fever, chills, cough, aches, vomit, and age_cat).  

Within the function itself, we provide the `method = ` as `glm` (no quotes), the `y = ` outcome column (`outcome`), specify to `method.args = ` that we want to run logistic regression via `family = binomial`, and we tell it to exponentiate the results.  

The output is HTML and contains the counts

```{r odds_gt, message=F, warning=F}

univ_tab <- linelist %>% 
  dplyr::select(explanatory_vars, outcome) %>% ## select variables of interest

  tbl_uvregression(                         ## produce univariate table
    method = glm,                           ## define regression want to run (generalised linear model)
    y = outcome,                            ## define outcome variable
    method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
    exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
  )

## view univariate results table 
univ_tab
```

There are many modifications you can make to this table output, such as adjusting the text labels, bolding rows by their p-value, etc. See tutorials [here](http://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html) and elsewhere online.  



<!-- ======================================================= -->

## Stratified {  }

Stratified analysis is currently still being worked on for **gtsummary**, 
this page will be updated in due course. 


<!-- ======================================================= -->

### **gtsummary** package {-}

TODO

<!-- ======================================================= -->

### **base** R {-}

TODO

<!-- ======================================================= -->



## Multivariable  

For multivariable analysis, we again present two approaches:  

* `glm()` and `tidy()`  
* **gtsummary** package  

The workflow is similar for each, as shown below, and only the last step of pulling a final table together is different.


### Conduct multivariable {-}  

Use `glm()` but add more variables to the right side of the equation, separated by plus symbols (`+`). 


To run the model with all of our explanatory variables we would run:  

```{r}
mv_reg <- glm(outcome ~ gender + fever + chills + cough + aches + vomit + age_cat, family = "binomial", data = linelist)

summary(mv_reg)
```

If you want to include two variables and an interaction between them you can separate them with an asterisk `*` instead of a `+`. Separate them with a colon `:` if you are only specifying the interaction. For example:  

```{r, eval=F}
glm(outcome ~ gender + age_cat * fever, family = "binomial", data = linelist)
```


*Optionally*, you can leverage the pre-defined vector of column names and re-create the above command using `str_c()` as shown below. This might be useful if your explanatory variable names are changing, or you don't want to type them all out again.  

```{r mv_regression}

## run a regression with all variables of interest 
mv_reg <- explanatory_vars %>%  ## begin with vector of explanatory column names
  str_c(collapse = "+") %>%     ## combine all names of the variables of interest separated by a plus
  str_c("outcome ~ ", .) %>%    ## combine the names of variables of interest with outcome in formula style
  glm(family = "binomial",      ## define type of glm as logistic,
      data = linelist)          ## define your dataset
```


#### Building the model {-}  

You can build you model step-by-step, saving various models that include certain explanatory variables. You can compare these models with likelihood-ratio tests using `lrtest()` from the package **lmtest**, as below:  

<span style="color: black;">**_NOTE:_** Using **base** `anova(model1, model2, test = "Chisq)` produces the same results </span> 

```{r}
model1 <- glm(outcome ~ age_cat, family = "binomial", data = linelist)
model2 <- glm(outcome ~ age_cat + gender, family = "binomial", data = linelist)

lmtest::lrtest(model1, model2)
```

Another option is to take the model object and apply the `step()` function from the **stats** package. Specify which variable selection direction you want use when building the model.      

```{r}
## choose a model using forward selection based on AIC
## you can also do "backward" or "both" by adjusting the direction
final_mv_reg <- mv_reg %>%
  step(direction = "forward", trace = FALSE)
```


You can also turn off scientific notation in your R session, for clarity:  

```{r}
options(scipen=999)
```

As described in the section on univariate analysis, pass the model output to `tidy()` to exponentiate the log odds and CIs. Finally we round all numeric columns to two decimal places. Scroll through to see all the rows.  

```{r mv_regression_base}

mv_tab_base <- final_mv_reg %>% 
  ## get a tidy dataframe of estimates 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
  ## round 
  mutate(across(is.numeric, round, digits = 2))
```

Here is what the resulting data frame looks like: 

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(mv_tab_base, rownames = FALSE, options = list(pageLength = 10, scrollX=T), class = 'white-space: nowrap' )
```





<!-- ======================================================= -->

### Combine univariate and multivariable {-}

#### Combine with **gtsummary**  {-}  

The `gtsummary` package provides the `tbl_regression` function, which will 
take the outputs from a regression (`glm()` in this case) and produce an easy 
summary table. You can also combine several different output tables produced by `gtsummary` with 
the `tbl_merge` function. 

```{r mv_regression_gt}
## show results table of final regression 
mv_tab <- tbl_regression(final_mv_reg, exponentiate = TRUE)
```

And now combine the univariate and multivariable results:  

```{r}
## combine with univariate results 
tbl_merge(
  tbls = list(univ_tab, mv_tab), 
  tab_spanner = c("**Univariate**", "**Multivariable**")) # set header names
```



#### Combine with **dplyr** {-}  

An alternative way of combining the `glm()`/`tidy()` univariate and multivariable outputs is with the **dplyr** join functions.  

* Join the univariate results from earlier (which contains counts) with the tidied multivariable results  
* Use `select()` to keep only the columns we want, specify their order, and re-name them  
* Use `round()` with two decimal places on all the column that are class Double  

```{r, warning=F, message=F}
## combine univariate and multivariable tables 
left_join(univ_tab_base, mv_tab_base, by = "term") %>% 
  ## choose columns and rename them
  select( # new name =  old name
    "characteristic" = term, 
    "recovered"      = "0", 
    "dead"           = "1", 
    "univ_or"        = estimate.x, 
    "univ_ci_low"    = conf.low.x, 
    "univ_ci_high"   = conf.high.x,
    "univ_pval"      = p.value.x, 
    "mv_or"          = estimate.y, 
    "mvv_ci_low"     = conf.low.y, 
    "mv_ci_high"     = conf.high.y,
    "mv_pval"        = p.value.y 
  ) %>% 
  mutate(across(is.double, round, 2))

```




<!-- ======================================================= -->

## Forest plot {  }

This section shows how to produce a plot with the outputs of your regression.
There are two options, you can build a plot yourself using `ggplot2` or use a 
meta-package called **easystats** (package that includes many packages).  


<!-- ======================================================= -->

### **ggplot2** package {-}

You can build a forest plot with `ggplot()` by plotting elements of the multivariable regression results. Add the layers:  

* estimates with `geom_point()`  
* confidence intervals with `geom_errorbar()`  
* a vertical line at OR = 1 with `geom_vline()`  

You may want to re-arrange the order of the variables/levels on the y-axis (see how the order of age_cat levels is alphabetical and not sensical). To do this, use `fct_relevel()` from the **forcats** package to classify the column `term` as a factor and specify the order manually. See the page on [Factors] for more details.  

```{r ggplot_forest}

## remove the intercept term from your multivariable results
mv_tab_base %>% 
  filter(term != "(Intercept)") %>% 
  ## plot with variable on the y axis and estimate (OR) on the x axis
  ggplot(aes(x = estimate, y = term)) +
  ## show the estimate as a point
  geom_point() + 
  ## add in an error bar for the confidence intervals
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high)) + 
  ## show where OR = 1 is for reference as a dashed line
  geom_vline(xintercept = 1, linetype = "dashed")
  
```


<!-- ======================================================= -->

### **easystats** packages {-}

The alternative if you do not want to decide all of the different things required
for a `ggplot`, is to use a combination of **easystats** packages. 
In this case the **parameters** package function `model_parameters()` does the equivalent
of **broom** package function `tidy()`. The **see** package then accepts those outputs
and creates a default forest plot as a `ggplot()` object. 

```{r easystats_forest}

## remove the intercept term from your multivariable results
final_mv_reg %>% 
  model_parameters(exponentiate = TRUE) %>% 
  plot()
  
```


<!-- ======================================================= -->

## Resources {  }

Much of the information in this page is adapted from these resources and vignettes online:  

[gtsummary](http://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html)  

[UCLA stats page](https://stats.idre.ucla.edu/other/dae/)  

[sthda stepwise regression](http://www.sthda.com/english/articles/36-classification-methods-essentials/150-stepwise-logistic-regression-essentials-in-r/)   

