
# Tables for presentation { }  

This section demonstrates how to create publication-ready tables, which can be inserted directly into shareable documents, including R Markdown outputs.  

See the [Descriptive tables] page for instruction on how to make tabulations, cross-tabulations, and descriptive summary tables.  

<!-- ======================================================= -->
## Overview {  }

We build on previous sections on basic statistics and creating summary tables (e.g. using **dplyr** and **gtsummary** and show how to create publication-read tables. The primary package we use is **flextable**, which is compatible with multiple R Markdown formats, including html and word documents. 

Example: 

**Table of Ebola patients with outcome information: Number, proportion, and CT values of cases who recovered and died** (with Military hospital highlighted)

```{r echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width=c('50%', '50%')}

linelist <- rio::import(here::here("data", "case_linelists", "linelist_cleaned.rds")) 

border_style = officer::fp_border(color="black", width=1)

pacman::p_load(
  rio,            # import/export
  here,           # file pathways
  flextable,      # make HTML tables 
  officer,        # helper functions for tables
  tidyverse)      # data management, summary, and visualization

table <- linelist %>% 
  # filter
  ########
  filter(!is.na(outcome) & hospital != "Missing") %>%  # Remove cases with missing outcome or hospital
  
  # Get summary values per hospital-outcome group
  ###############################################
  group_by(hospital, outcome) %>%                      # Group data
  summarise(                                           # Create new summary columns of indicators of interest
    N = n(),                                            # Number of rows per hospital-outcome group     
    ct_value = median(ct_blood, na.rm=T)) %>%           # median CT value per group
  
  # add totals
  ############
  bind_rows(                                           # Bind the previous table with this mini-table of totals
    linelist %>% 
      filter(!is.na(outcome) & hospital != "Missing") %>%
      group_by(outcome) %>%                            # Grouped only by outcome, not by hospital    
      summarise(
        N = n(),                                       # Number of rows for whole dataset     
        ct_value = median(ct_blood, na.rm=T))) %>%     # Median CT for whole dataset
  
  # Pivot wider and format
  ########################
  mutate(hospital = replace_na(hospital, "Total")) %>% 
  pivot_wider(                                         # Pivot from long to wide
    values_from = c(ct_value, N),                       # new values are from ct and count columns
    names_from = outcome) %>%                           # new column names are from outcomes
  mutate(                                              # Add new columns
    N_Known = N_Death + N_Recover,                      # number with known outcome
    Pct_Death = N_Death / N_Known * 100,               # percent cases who died
    Pct_Recover = N_Recover/N_Known * 100) %>%         # percent who recovered
  select(                                              # Re-order columns
    hospital, N_Known,                                   # Intro columns
    N_Recover, Pct_Recover, ct_value_Recover,            # Recovered columns
    N_Death, Pct_Death, ct_value_Death)  %>%             # Death columns
  arrange(N_Known) %>%                                 # Arrange rows from lowest to highest (Total row at bottom)

  # formatting
  ############
  flextable() %>% 
  add_header_row(
    top = TRUE,                # New header goes on top of existing header row
    values = c("Hospital",     # Header values for each column below
               "Total cases with known outcome", 
               "Recovered",    # This will be the top-level header for this and two next columns
               "",
               "",
               "Died",         # This will be the top-level header for this and two next columns
               "",             # Leave blank, as it will be merged with "Died"
               "")) %>% 
    set_header_labels(         # Rename the columns in original header row
      hospital = "", 
      N_Known = "",                  
      N_Recover = "Total",
      Pct_Recover = "% of cases",
      ct_value_Recover = "Median CT values",
      N_Death = "Total",
      Pct_Death = "% of cases",
      ct_value_Death = "Median CT values")  %>% 
  merge_at(i = 1, j = 3:5, part = "header") %>% # Horizontally merge columns 3 to 5 in new header row
  merge_at(i = 1, j = 6:8, part = "header") %>%  
  border_remove() %>%  
  theme_booktabs() %>% 
  vline(part = "all", j = 2, border = border_style) %>%   # at column 2 
  vline(part = "all", j = 5, border = border_style) %>%   # at column 5
  merge_at(i = 1:2, j = 1, part = "header") %>% 
  merge_at(i = 1:2, j = 2, part = "header") %>% 
  width(j=1, width = 2.7) %>% 
  width(j=2, width = 1.5) %>% 
  width(j=c(4,5,7,8), width = 1) %>% 
  flextable::align(., align = "center", j = c(2:8), part = "all") %>% 
  bg(., part = "body", bg = "gray95")  %>% 
  bg(., j=c(1:8), i= ~ hospital == "Military Hospital", part = "body", bg = "#91c293") %>% 
  colformat_num(., j = c(4,7), digits = 1) %>%
  bold(i = 1, bold = TRUE, part = "header") %>% 
  bold(i = 6, bold = TRUE, part = "body")

table
```


<!-- ======================================================= -->
## Preparation {  }

### Load packages {-} 

Load, and install if necessary, **flextable**, which we will use to convert the above table into a fully formatted and presentable table. In this handbook we emphasize `p_load()` from **pacman**, which installs the package if necessary *and* loads it for use. You can also load packages with `library()` from **base** R. See the page on [R basics] for more information on R packages.  

```{r}
pacman::p_load(
  rio,            # import/export
  here,           # file pathways
  flextable,      # make HTML tables 
  officer,        # helper functions for tables
  tidyverse)      # data management, summary, and visualization

```

### Import data {-}  

We import the dataset of cases from a simulated Ebola epidemic. If you want to download the data to follow step-by-step, see instructions in the [Download handbook and data] page. The dataset is imported using the `import()` function from the **rio** package. See the page on [Import and export] for various ways to import data.

```{r, echo=F}
# import the linelist into R
linelist <- rio::import(here::here("data", "case_linelists", "linelist_cleaned.rds"))
```

```{r, eval=F}
# import the linelist
linelist <- import("linelist_cleaned.xlsx")
```

The first 50 rows of the linelist are displayed below.

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist, 50), rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap' )
```

### Prepare table {-}  

*Before* beginning to use **flextable** you will need to *create* your table contents. Use packages discussed in other sections such as **janitor** or **dplyr** to create a table with the content of interest, with the correct columns and rows. 
  
Below is an example - we create a simple summary table of patient outcomes. We are interested in knowing the number and proportion of patients that recover or died, as well as their median CT values, by hospital of admission. 

```{r message=FALSE, warning=FALSE}
table <- linelist %>% 
  # filter
  ########
  filter(!is.na(outcome) & hospital != "Missing") %>%  # Remove cases with missing outcome or hospital
  
  # Get summary values per hospital-outcome group
  ###############################################
  group_by(hospital, outcome) %>%                      # Group data
  summarise(                                           # Create new summary columns of indicators of interest
    N = n(),                                            # Number of rows per hospital-outcome group     
    ct_value = median(ct_blood, na.rm=T)) %>%           # median CT value per group
  
  # add totals
  ############
  bind_rows(                                           # Bind the previous table with this mini-table of totals
    linelist %>% 
      filter(!is.na(outcome) & hospital != "Missing") %>%
      group_by(outcome) %>%                            # Grouped only by outcome, not by hospital    
      summarise(
        N = n(),                                       # Number of rows for whole dataset     
        ct_value = median(ct_blood, na.rm=T))) %>%     # Median CT for whole dataset
  
  # Pivot wider and format
  ########################
  mutate(hospital = replace_na(hospital, "Total")) %>% 
  pivot_wider(                                         # Pivot from long to wide
    values_from = c(ct_value, N),                       # new values are from ct and count columns
    names_from = outcome) %>%                           # new column names are from outcomes
  mutate(                                              # Add new columns
    N_Known = N_Death + N_Recover,                      # number with known outcome
    Pct_Death = N_Death / N_Known * 100,               # percent cases who died
    Pct_Recover = N_Recover/N_Known * 100) %>%         # percent who recovered
  select(                                              # Re-order columns
    hospital, N_Known,                                   # Intro columns
    N_Recover, Pct_Recover, ct_value_Recover,            # Recovered columns
    N_Death, Pct_Death, ct_value_Death)  %>%             # Death columns
  arrange(N_Known)                                    # Arrange rows from lowest to highest (Total row at bottom)

table  # print

```




<!-- ======================================================= -->
## Basic flextable {  }

**Creating a flextable**

To create and manage **flextable** objects, we pass the table object through the `flextable()` function and progressively pipe the object through more **flextable** formatting functions. The general syntax of each line of flextable code is as follows:

* `function(table, i = X, j = X, part = "X")`, where:
  * `table = ` is the name of the table object, although does not need to be stated if the table is piped into the function.
  * The 'function' can be one of many different functions, such as `width()` to determine column widths, `bg()` to set background colours, `align()` to set whether text is centre/right/left aligned, and so on. 
  * `part = ` refers to which part of the table the function is being applied to. E.g. "header", "body" or "all". 
  * `i = ` specifies the *row* to apply the function to, where 'X' is the row number. If multiple rows, e.g. the first to third rows, one can specify: `i = c(1:3)`. Note if 'body' is selected, the first row starts from underneath the header section.
  * `j = ` specifies the *column* to apply the function to, where 'x' is the column number or name. If multiple columns, e.g. the fifth and sixth, one can specify: `j = c(5,6)`. 
  
You can find the complete list of **flextable** formatting function [here](https://davidgohel.github.io/flextable/reference/index.html) or review the documentation with `?flextable`.  


```{r}

my_table <- flextable(table) 
my_table

```


**Formatting cell content**

We can ensure that the proportion columns display only one decimal place using the function `colformat_num()`. Note this could also have been done at data management stage with the `round()` function. 

```{r}
my_table <- colformat_num(my_table, j = c(4,7), digits = 1)
my_table
```

**Formatting column width**

We can use the `autofit()` function, which nicely stretches out the table so that each cell only has one row of text. The function `qflextable()` is a convenient shorthand for `flextable()` and `autofit()`.  

```{r}

my_table %>% autofit()

```

However, this might not always be appropriate, especially if there are very long values within cells, meaning the table might not fit on the page. 

Instead, we can specify widths with the `width()` function. It can take some playing around to know what width value to put. In the example below, we specify different widths for column 1, column 2, and columns 4 to 8. 

```{r}

my_table <- my_table %>% 
  width(j=1, width = 2.7) %>% 
  width(j=2, width = 1.5) %>% 
  width(j=c(4,5,7,8), width = 1)

my_table
  
```

**Column headers**

We want to clearer headers for easier interpretation of table contents.

First we can add an extra header layer for clarity. We do this with the `add_header_row()` function with `top = ` set to TRUE, so that columns covering the same subgroups can be grouped together. We provide the new name of each column to `values = `, leaving empty values "" for column we know we will merge together later.  

We also rename the now-second header. Finally we use `merge_at()` to merge the column headers in the top header row. 

```{r}
my_table <- my_table %>% 
  add_header_row(
    top = TRUE,                # New header goes on top of existing header row
    values = c("Hospital",     # Header values for each column below
               "Total cases with known outcome", 
               "Recovered",    # This will be the top-level header for this and two next columns
               "",
               "",
               "Died",         # This will be the top-level header for this and two next columns
               "",             # Leave blank, as it will be merged with "Died"
               "")) %>% 
    set_header_labels(         # Rename the columns in original header row
      hospital = "", 
      N_Known = "",                  
      N_Recover = "Total",
      Pct_Recover = "% of cases",
      ct_value_Recover = "Median CT values",
      N_Death = "Total",
      Pct_Death = "% of cases",
      ct_value_Death = "Median CT values")  %>% 
  merge_at(i = 1, j = 3:5, part = "header") %>% # Horizontally merge columns 3 to 5 in new header row
  merge_at(i = 1, j = 6:8, part = "header")     # Horizontally merge columns 6 to 8 in new header row

my_table  # print

```

**Formatting borders and background **

You can adjust the borders, internal lines, etc. with various **flextable** functions. It is often easier to start by removing all existing borders with `border_remove()`.  

Then, you can apply default border themes by passing the table to `theme_box()`, `theme_booktabs()`, or `theme_alafoli()`.  

You can add vertical and horizontal lines with a variety of functions. `hline()` and `vline()` add lines to a specified row or column, respectively. Within each, you must specify the `part = ` as either "all", "body", or "header". For vertical lines, specify to `j = ` the column, and for horizontal line to `i = `. Other functions like `vline_right()`, `vline_left()`, `hline_top()`, and `hline_bottom()` add lines to the outsides only.  

In all of these functions, the actual line style itself must be specified to `border = ` and must be the output of the `fp_border()` function from the **officer** package. This function helps you define the width and color or the line. You can define this outside of the table, as shown below.  

```{r}
# define style for border line
border_style = officer::fp_border(color="black", width=1)

# add border lines to table
my_table <- my_table %>% 

  # Remove all existing borders
  border_remove() %>%  
  
  # add horizontal lines via a pre-determined theme setting
  theme_booktabs() %>% 
  
  # add vertical lines to separate Recovered and Died sections
  vline(part = "all", j = 2, border = border_style) %>%   # at column 2 
  vline(part = "all", j = 5, border = border_style)       # at column 5

my_table
```

**Font and alignment**

We centre-align all columns aside from the left-most column with the hospital names, using the `align()` function from **flextable**.

```{r}
my_table <- my_table %>% 
   flextable::align(align = "center", j = c(2:8), part = "all") 
my_table
```

Additionally, we can increase the header font size and change then to bold. We can also change the total row to bold.  

```{r}

my_table <-  my_table %>%  
  fontsize(i = 1, size = 12, part = "header") %>%   # adjust font size of header
  bold(i = 1, bold = TRUE, part = "header") %>%     # adjust bold face of header
  bold(i = 6, bold = TRUE, part = "body")           # adjust bold face of total row (row 6)

my_table

```

**Merged cells**  

Just as we merge cells horizontally in the header row, we can also merge cells vertically using `merge_at()` and specifying the rows (`i`) and column (`j`).  

```{r}
my_table <- my_table %>% 
  merge_at(i = 1:2, j = 1, part = "header") %>% 
  merge_at(i = 1:2, j = 2, part = "header")

```

**Background**

To distinguish the content of the table from the headers, we may want to add additional formatting. e.g. changing the background colour. In this example we change the table body to gray.

```{r}
my_table <- my_table %>% 
    bg(part = "body", bg = "gray95")  

my_table 
```


<!-- ======================================================= -->
## Conditional flextable formatting {  }

We can  highlight all values in a column that meet a certain rule, e.g. where more than 55% of cases died. 

```{r}

my_table %>% 
  bg(j=7, i= ~ Pct_Death >=55, part = "body", bg = "red") 

```



Or, we can higlight the entire row meeting a certain criterion, such as a hospital of interest. This is particularly helpful when looping through e.g. reports per geographical area, to highlight in tables where the current iteration compares to the other geographies. To do this we just remove the column (`j`) specification.


```{r}

my_table %>% 
  bg(., j=c(1:8), i= ~ hospital == "Military Hospital", part = "body", bg = "#91c293") 

```

<!-- ======================================================= -->
## Saving your table {  }

There are different ways the table can be integrated into your output. 

**Save single table**

You can export the tables to Word, PowerPoint or HTML or as an image (PNG) files. To do this, one of the following functions is used:

* `save_as_docx()`  
* `save_as_pptx()`  
* `save_as_image()`  
* `save_as_html()`  

For instance: 
```{r message=FALSE, warning=FALSE, eval=F}
save_as_docx("my table" = my_table, path = "file.docx")
# Edit the 'my table' as needed for the title of table. If not specified the whole file will be blank. 

save_as_image(my_table, path = "file.png")
```

Note the packages `webshot` or `webshot2` are required to save a flextable as an image. Images may come out with transparent backgrounds.

If you want to view a 'live' versions of the flextable output in the intended document format, for instance so you can see if it fits in the page or so you can copy it into another document, you can use the print method with the argument preview set to “pptx” or “docx”. The document will pop up.

```{r, eval=F}
print(my_table, preview = "docx") # Word document example
print(my_table, preview = "pptx") # Powerpoint example
```

**Save table to R markdown document**

This table can be integrated into your an automated document, an R markdown output, if the table object is called within the R markdown chunk. This means the table can be updated as part of a report where the data might change, so the numbers can be refreshed.

See detail in the [R Markdown reports] page of this handbook. 

<!-- ======================================================= -->
## Resources {  }

The full **flextable** book is here: https://ardata-fr.github.io/flextable-book/
The Github site is [here](https://davidgohel.github.io/flextable/)  
A manual of all the **flextable** functions can be found [here](https://davidgohel.github.io/flextable/reference/index.html)

A gallery of beautiful example **flextable** tables with code can be accessed [here](https://ardata-fr.github.io/flextable-gallery/gallery/)  
