
# Tables { }  

This section demonstrates how to create publication-ready tables, which can be inserted directly into shareable documents, including R Markdown outputs.  

See the [Descriptive analysis] page for instruction on how to make tabulations, cross-tabulations, and descriptive summary tables.  

<!-- ======================================================= -->
## Overview {  }

We build on previous sections on basic statistics and creating summary tables (e.g. using `dplyr` and `gtsummary` and show how to create publication-read tables. The primary package we use is flextable, which is compatible with multiple R Markdown formats, including html and word documents. 

Example:

**Table of Ebola patients with outcome information: Number, proportion, and CT values of cases who recovered and died**

```{r echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width=c('50%', '50%')}

linelist <- rio::import(here::here("data", "linelist_cleaned.rds")) 

library(flextable)
library(dplyr)
library(tidyr)
library(officer)

linelist %>% 
  group_by(hospital, outcome) %>% 
  filter(!is.na(outcome) & hospital!="Missing") %>% 
  summarise(ct_value = median(ct_blood), N = n()) %>% 
  pivot_wider(values_from=c(ct_value, N), names_from = outcome) %>% 
  mutate(`N known` = `N_Death` + N_Recover) %>% 
  arrange(-`N known`) %>% 
  mutate(`Prop_Death` = `N_Death`/`N known`*100,
         `Prop_Recover` = `N_Recover`/`N known`*100) %>% 
  select(hospital, `N known`, `N_Recover`, `Prop_Recover`, ct_value_Recover,
         `N_Death`, `Prop_Death`, ct_value_Death) %>% 
  flextable() %>% 
  set_header_labels(hospital = "Hospital",
                    `N known` = "Total cases with known outcome",                  
                    N_Recover = "Recovered",
                    Prop_Recover = "",
                    ct_value_Recover = "",
                    N_Death = "Died",
                    Prop_Death = "",
                     ct_value_Death = "") %>% 
add_header_row(.,
   values = c("", "", "Number", "Proportion of cases with outcomes", "CT value",
              "Number", "Proportion of cases with outcomes", "CT value"), top = F) %>% 
  merge_at(., i = 1, j = 3:5, part = "header") %>% 
  merge_at(., i = 1, j = 6:8, part = "header") %>% 
  hline(., i=2, part = "header", border = fp_border(color="black", width=2))  %>% 
  hline(., i=1, part = "header", border = fp_border(color="white", width=2)) %>% 
  width(., j=1, width = 3) %>% 
  width(., j=2, width = 1.5) %>% 
  width(., j=c(4,7), width = 1.5) %>% 
  flextable::align(., align = "center", j = c(2:8), part = "all") %>% 
  bg(., part = "body", bg = "gray95")  %>% 
  bg(., j=c(1:8), i= ~ hospital == "Military Hospital", part = "body", bg = "#91c293") %>% 
  colformat_num(., j = c(4,7), digits = 2) %>% 
  bold(i = 1, bold = TRUE, part = "header")


```


<!-- ======================================================= -->
## Preparation {  }

### Load packages {-} 

Use packages discussed in other sections such as `gtsummary` and `dplyr` to create a table with the content of interest, with the correct columns and rows. 

Load, and install if necessary, `flextable`, which we will use to convert the above table into a fully formatted and presentable table. In this handbook we emphasize `p_load()` from **pacman**, which installs the package if necessary and loads it for use. You can also load packages with `library()` from **base** R. See the page on [R basics] for more information on R packages.  

```{r}
pacman::p_load(flextable)

```

### Import data {-}  


We import the dataset of cases from a simulated Ebola epidemic. If you want to download the data to follow step-by-step, see instructions in the [Datasets used] page. The dataset is imported using the `import()` function from the **rio** package. See the page on [Import and export] for various ways to import data.

```{r, echo=F}
# import the linelist into R
linelist <- rio::import(here::here("data", "linelist_cleaned.rds"))
```

```{r, eval=F}
# import the linelist
linelist <- import("linelist_cleaned.xlsx")
```

The first 50 rows of the linelist are displayed below.

```{r, message=FALSE, echo=F}
# display the linelist data as a table
DT::datatable(head(linelist, 50), rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap' )
```

### Prepare table {-}  

*Before* beginning to use **flextable** you will need to *create* your table contents.  

Below is an example - we create a simple summary table of patient outcomes. We are interested in knowing the number and proportion of patients that recover or died, as well as their median CT values, by hospital of admission. 

```{r message=FALSE, warning=FALSE}
table <- linelist %>% 
  group_by(hospital, outcome) %>%                      # Group data
  filter(!is.na(outcome) & hospital != "Missing") %>%  # Remove cases with missing outcome or hospital
  summarise(
    ct_value = median(ct_blood),                       # Create new summary columns of indicators of interest
    N = n()) %>%                                 
  pivot_wider(                                         # Pivot from long to wide
    values_from = c(ct_value, N),
    names_from = outcome) %>% 
  mutate(                                              # Calculate total number
    `N known` = `N_Death` + N_Recover) %>% 
  arrange(                                             # Arrange rows from highest to lowest total
    -`N known`) %>% 
  mutate(                                              # Calculate proportions
    `Prop_Death` = `N_Death`/`N known`*100,  
    `Prop_Recover` = `N_Recover`/`N known`*100) %>% 
  select(                                              # Re-order columns
    hospital, `N known`, `N_Recover`,
    `Prop_Recover`, ct_value_Recover,
    `N_Death`, `Prop_Death`, ct_value_Death)  

table  # print

```




<!-- ======================================================= -->
## Basic flextable {  }

**Creating a flextable**

To create and manage flextable objects, we pass the table object through the flextable function and progressively add more formatting and features using the dplyr syntax. 

The syntax of each line of flextable code is as follows:

* `function(table, i = X, j = X, part = "X")`, where:
  * `table` is the name of the table object, although does not need to be stated if using the dplyr syntax and the table name has already been specified (see examples).
  * The 'function' can be one of many different functions, such as `width` to determine column widths, `bg` to set background colours, `align` to set whether text is centre/right/left aligned, and so on. 
  * `part` refers to which part of the table the function is being applied to. E.g. "header", "body" or "all". 
  * `i` specifies the row to apply the function to, where 'X' is the row number. If multiple rows, e.g. the first to third rows, one can specify: `i = c(1:3)`. Note if 'body' is selected, the first row starts from underneath the header section.
  * `j` specifies the column to apply the function to, where 'x' is the column number or name. If multiple rows, e.g. the fifth and sixth, one can specify: `j = c(5,6)`. 


```{r}

ftable <- flextable(table) 
ftable

```

We see immediately that it has suboptimal spacing, and the proportions have too many decimal places. 

**Formatting cell content**

We edit the proportion colums to one decimal place using flextable code. Note this could also have been done at data management stage with the `round()` function. 

```{r}
ftable <- colformat_num(ftable, j = c(4,7), digits = 1)
ftable
```

**Formatting column width**

We can use the autofit() function, which nicely stretches out the table so that each cell only has one row of text. 

```{r}

ftable %>% autofit()

```

However, this might not always be appropriate, especially if there are very long values within cells, meaning the table might not fit on the page. 

Instead, we can specify widths. It can take some playing around to know what width value to put. In the example below, we specify different widths for column 1, column 2, and columns 4 to 8. 

```{r}

ftable <- ftable %>% 
  width(j=1, width = 2.7) %>% 
  width(j=2, width = 1.5) %>% 
  width(j=c(4,5,7,8), width = 1)

ftable
  
```

**Column headers**

We want to clearer headers for easier interpretation of table contents.

First we can add an extra header layer for clarity. We do this with the `add_header_row` with 'top' set to true, so that columns about the same subgroups can be grouped together. We also rename the now-second header layer. Finally we merge the columns in the top header row. 

```{r}
ftable <- ftable %>% 
  add_header_row( values = c("Hospital", 
                             "Total cases with known outcome", 
                             "Recovered", 
                             "",
                             "",
                             "Died", # This and the next two columns will be one value ' died
                             "",     # As merging will keep the first column, next two are blank
                             ""), 
                  top = T) %>% # New header goes on top of existing header row
    set_header_labels(hospital = "", # Rename the columns in original header row
                    `N known` = "",                  
                    N_Recover = "Total",
                    Prop_Recover = "% of cases",
                    ct_value_Recover = "Median CT values",
                    N_Death = "Total",
                    Prop_Death = "% of cases",
                     ct_value_Death = "Median CT values")  %>% 
  merge_at(i = 1, j = 3:5, part = "header") %>% # Horizontally merge columns in new header row
  merge_at(i = 1, j = 6:8, part = "header")  

ftable

```

**Formatting borders and background **

Flextable has default borders that do not respond well to additional header levels. We start from scratch by removing the existing borders with `border_remove`. Then we add a black line to the bottom of the table using `hlinw`, by specifying the 5th row of the table body. Flextable will default add a line to the bottom of the row. In order to add black lines to the _top_ of sections, we need to use `hline_top`. 

We also use `fp_border` here, which actually applied the border. This is a function from the `officer` package.

```{r}
  
library(officer)

ftable <- ftable %>% 
  border_remove() %>% # Remove existing borders 
  hline(part = "body", i=5, border = fp_border(color="black", width=2)) %>% 
  hline_top(part = "header", border = fp_border(color="black", width=2)) %>%
  hline_top(part = "body", border = fp_border(color="black", width=2)) 


ftable

```

**Font and alignment**

We centre-align all columns aside from the left-most column with the hospital names, using the `align` function.

```{r}
ftable <- ftable %>% 
   flextable::align(align = "center", j = c(2:8), part = "all") 
ftable
```

Additionally, we can increase the header font size and change then to bold. 

```{r}

ftable <-  ftable %>%  
  fontsize(i = 1, size = 12, part = "header") %>% 
    bold(i = 1, bold = TRUE, part = "header")

ftable

```

**Background**

To distinguish the content of the table from the headers, we may want to add additional formatting. e.g. changing the background colour. In this example we change the table body to gray.

```{r}
ftable <- ftable %>% 
    bg(part = "body", bg = "gray95")  

ftable 
```


<!-- ======================================================= -->
## Conditional flextable formatting {  }

We can  highlight all values in a column that meet a certain rule, e.g. where more than 55% of cases died. 

```{r}

ftable %>% 
  bg(j=7, i= ~ Prop_Death >=55, part = "body", bg = "red") 

```



Or, we can higlight the entire row meeting a certain criterion, such as a hospital of interest. This is particularly helpful when looping through e.g. reports per geographical area, to highlight in tables where the current iteration compares to the other geographies. To do this we just remove the column (`j`) specification.


```{r}

ftable %>% 
  bg(., j=c(1:8), i= ~ hospital == "Military Hospital", part = "body", bg = "#91c293") 

```

<!-- ======================================================= -->
## Saving your table {  }

There are different ways the table can be integrated into your output. 

**Save single table**

You can export the tables to Word, PowerPoint or HTML or as an image (PNG) files. To do this, one of the following functions is used:

* save_as_docx
* save_as_pptx
* save_as_image
* save_as_html

For instance: 
```{r message=FALSE, warning=FALSE}
save_as_docx("my table" = ftable, path = "file.docx")
# Edit the 'my table' as needed for the title of table. If not specified the whole file will be blank. 

save_as_image(ftable, path = "file.png")
```

Note the packages `webshot` or `webshot2` are required to save a flextable as an image.Images may come out with transparent backgrounds.

If you want to view a 'live' versions of the flextable output in the intended document format, for instance so you can see if it fits in the page or so you can copy it into another document, you can use the print method with the argument preview set to “pptx” or “docx”. The document will pop up.

```
print(ftable, preview = "docx") # Word document example
print(ftable, preview = "pptx") # Powerpoint example
```

**Save table to R markdown document**

This table can be integrated into your an automated document, an R markdown output, if the table object is called within the R markdown chunk. This means the table can be updated as part of a report where the data might change, so the numbers can be refreshed.

See detail in the R markdown section of this handbook. 

<!-- ======================================================= -->
## Resources {  }

The full flextable explanation is here: https://ardata-fr.github.io/flextable-book/

