
# Time series analysis {.tabset .tabset-fade}  

<!-- ======================================================= -->
## Overview {.tabset .tabset-fade .tabset-pills}

This tab demonstrates the use of several packages for time series analysis. 
It primarily relies on packages from the [**tidyverts**](https://tidyverts.org/) 
family, but will also use the RECON [**trending**](https://github.com/reconhub/trending) 
package to fit models that are more appropriate for infectious disease epidemiology. 

1.  Time series data 
2.  Descriptive analysis 
3.  Fitting regressions
4.  Relation of two time series 
5.  Outbreak detection
6.  Interrupted time series


<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}

### Packages

This code chunk shows the loading of packages required for the analyses.

```{r load_packages}
pacman::p_load(rio,          # File import
               here,         # File locator
               tidyverse,    # data management + ggplot2 graphics
               tsibble,      # handle time series datasets
               slider,       # for calculating moving averages
               imputeTS,     # for filling in missing values
               feasts,       # for time series decomposition and autocorrelation
               forecast,     # fit sin and cosin terms to data (note: must load after fable)
               trending,     # fit and assess models 
               tmaptools,    # for getting geocoordinates (lon/lat) based on place names
               ecmwfr,       # for interacting with copernicus sateliate CDS API
               stars,        # for reading in .nc (climate data) files
               units         # for defining units of measurement (climate data)
               )
``` 

### Load data

The example dataset used in this section:

-   Weekly counts of campylobacter cases reported in Germany between 2001 and 2011. 

This dataset is a reduced version of the dataset available in the [**surveillance**](https://cran.r-project.org/web/packages/surveillance/) package. 
(for details load the surveillance package and see `?campyDE`)

The dataset is imported using the `import()` function from the *rio* package. See the *page on importing data* for various ways to import data.

```{r read_data_hide, echo=F}
# import the counts into R
counts <- rio::import(here::here("data", "campylobacter_germany.xlsx"))

```

```{r read_data_show, eval=F}
# import the linelist
counts <- rio::import("campylobacter_germany.xlsx")
```

The first 10 rows of the counts are displayed below.

```{r inspect_data, message=FALSE, echo=F}
# display the counts data as a table
DT::datatable(head(counts, 10), rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
```

### Clean data

The below makes sure that the date column is in the appropriate format. 
For this tab we will be using the **tsibble** package and so the `yearweek` 
function will be used to create a calendar week variable. There are several other
ways of doing this (see **handling dates page** for details), however for 
time series its best to keep within one framework. 

```{r clean_data}

## ensure the date column is in the appropriate format
counts$date <- as.Date(counts$date)

## create a calendar week variable 
## fitting ISO definitons of weeks starting on a monday
counts <- counts %>% 
     mutate(epiweek = yearweek(date, week_start = 1))

```

### Download climate data 

In the *relation of two time series* section of this tab, we will be comparing 
campylobacter case counts to climate data. 

Climate data for anywhere in the world can be downloaded from the EU's Copernicus 
Satellite. These are not exact measurements, but based on a model (similar to 
interpolation), however the benefit is global hourly coverage as well as forecasts. 

We will be using the **ecmwfr** package to pull data from the Copernicus 
climate data store. You will need to create a free account in order for this to 
work. The package website has a useful [walkthrough](https://github.com/bluegreen-labs/ecmwfr#use-copernicus-climate-data-store-cds)
of how to do this. Below is example code of how to go about doing this, once you 
have the appropriate API keys. You have to replace the X's below with your account
IDs.
You will need to download one year of data at a time otherwise the server times-out. 

If you are not sure of the coordinates for a location you want to download data 
for, you can use the **tmaptools** package to pull the coordinates off open street
maps. An alternative option is the [**photon**](https://github.com/rCarto/photon)
package, however this has not been released on to CRAN yet; the nice thing about 
**photon** is that it provides more contextual data for when there are several 
matches for your search.

```{r weather_data, eval = FALSE}

## retrieve location coordinates
coords <- geocode_OSM("Germany", geometry = "point")

## pull together long/lats in format for ERA-5 querying (bounding box) 
## (as just want a single point can repeat coords)
request_coords <- str_glue_data(coords$coords, "{y}/{x}/{y}/{x}")


## Pulling data modelled from copernicus satellite (ERA-5 reanalysis)
## https://cds.climate.copernicus.eu/cdsapp#!/software/app-era5-explorer?tab=app
## https://github.com/bluegreen-labs/ecmwfr

## set up key for weather data 
wf_set_key(user = "XXXXX",
           key = "XXXXXXXXX-XXXX-XXXX-XXXX-XXXXXXXXXXX",
           service = "cds") 

## run for each year of interest (otherwise server times out)
for (i in 2002:2011) {
  
  ## pull together a query 
  ## see here for how to do: https://bluegreen-labs.github.io/ecmwfr/articles/cds_vignette.html#the-request-syntax
  ## change request to a list using addin button above (python to list)
  ## Target is the name of the output file!!
  request <- request <- list(
    product_type = "reanalysis",
    format = "netcdf",
    variable = c("2m_temperature", "total_precipitation"),
    year = c(i),
    month = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"),
    day = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12",
            "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24",
            "25", "26", "27", "28", "29", "30", "31"),
    time = c("00:00", "01:00", "02:00", "03:00", "04:00", "05:00", "06:00", "07:00",
             "08:00", "09:00", "10:00", "11:00", "12:00", "13:00", "14:00", "15:00",
             "16:00", "17:00", "18:00", "19:00", "20:00", "21:00", "22:00", "23:00"),
    area = request_coords,
    dataset_short_name = "reanalysis-era5-single-levels",
    target = paste0("germany_weather", i, ".nc")
  )
  
  ## download the file and store it in the current working directory
  file <- wf_request(user     = "XXXXX",  # user ID (for authentication)
                     request  = request,  # the request
                     transfer = TRUE,     # download the file
                     path     = here::here("data", "Weather")) ## path to save the data
  }

```

### Load climate data

```{r}

## define path to weather folder 
file_paths <- list.files(
  here::here("data", "weather"), 
  full.names = TRUE)

## only keep those with the current name of interest 
file_paths <- file_paths[str_detect(file_paths, "germany")]


## read in as a stars object 
data <- stars::read_stars(file_paths)

## change to a data frame 
temp_data <- as_tibble(data) %>% 
  ## add in variables and correct units
  mutate(
    ## create an calendar week variable 
    epiweek = tsibble::yearweek(time), 
    ## create a date variable (start of calendar week)
    date = as.Date(epiweek),
    ## change temperature from kelvin to celsius
    t2m = set_units(t2m, celsius), 
    ## change precipitation from metres to millimetres 
    tp  = set_units(tp, mm)) %>% 
  ## group by week (keep the date too though)
  group_by(epiweek, date) %>% 
  ## get the average per week
  summarise(t2m = as.numeric(mean(t2m)), 
            tp = as.numeric(mean(tp)))

```




<!-- ======================================================= -->
## Time series data {.tabset .tabset-fade .tabset-pills}

There are a number of different packages for structuring and handling time series
data. As said, we will focus on the **tidyverts** family of packages and so will
use the **tsibble** package to define our time series object. Having a data set
defined as a time series object means it is much easier to structure our analysis. 

To do this we use the `tsibble` function and specify the "index", i.e. the variable
specifying the time unit of interest. In our case this is the *epiweek* variable. 

If we had a data set with weekly counts by province, for example, we would also 
be able to specify the grouping variable using the "key" argument. 
This would allow us to do analysis for each group. 


```{r ts_object}

## define time series object 
counts <- tsibble(counts, index = epiweek)

```

Looking at `class(counts)` tells you that on top of being a tidy data frame 
("tbl_df", "tbl", "data.frame"), it has the additional properties of a time series
data frame ("tbl_ts"). 

You can take a quick look at your data by using ggplot2. We see from the plot that
there is a clear seasonal pattern, and that there are no missings. However, there
seems to be an issue with reporting at the beginning of each year; cases drop 
in the last week of the year and then increase for the first week of the next year. 

```{r basic_plot}

## plot a line graph of cases by week
ggplot(counts, aes(x = epiweek, y = case)) + 
     geom_line()

```


<span style="color: red;">**_DANGER:_** Most datasets aren't as clean as this example. 
You will need to check for duplicates and missings as below. </span>

<!-- ======================================================= -->
### duplicates {.tabset .tabset-fade .tabset-pills}

**tsibble** does not allow duplicate observations. So each row will need to be
unique, or unique within the group (`key` variable). 
The package has a few functions that help to identify duplicates. These include
`are_duplicated` which gives you a TRUE/FALSE vector of whether the row is a 
duplicate, and `duplicates` which gives you a data frame of the duplicated rows. 

See **de-duplication page** for details of how to select rows you want. 

```{r duplicates, eval = FALSE}

## get a vector of TRUE/FALSE whether rows are duplicates
are_duplicated(counts, index = epiweek) 

## get a data frame of any duplicated rows 
duplicates(counts, index = epiweek) 

```

<!-- ======================================================= -->
### Missings {.tabset .tabset-fade .tabset-pills}

We saw from our brief inspection above that there are no missings, but we also 
saw there seems to be a problem with reporting delay around new year. 
One way to address this problem could be to set these values to missing and then 
to impute values. The simplest form of time series imputation is to draw
a straight line between the last non-missing and the next non-missing value. 
To do this we will use the **imputeTS** package function `na_interpolation`. 
See **missing data page** for other options for imputation.  
Another alternative would be to calculating a moving average, to try and smooth
over these apparent reporting issues (see next section). 

```{r missings}

## create a variable with missings instead of weeks with reporting issues
counts <- counts %>% 
     mutate(case_miss = if_else(
          ## if epiweek contains 52, 53, 1 or 2
          str_detect(epiweek, "W51|W52|W53|W01|W02"), 
          ## then set to missing 
          NA_real_, 
          ## otherwise keep the value in case
          case
     ))

## alternatively interpolate missings by linear trend 
## between two nearest adjacent points
counts <- counts %>% 
  mutate(case_int = na_interpolation(case_miss)
         )

## to check what values have been imputed compared to the original
plotNA.imputations(counts$case_miss, counts$case_int)

```




<!-- ======================================================= -->
## Descriptive analysis {.tabset .tabset-fade .tabset-pills}



<!-- ======================================================= -->
### Moving averages {.tabset .tabset-fade .tabset-pills}

If data is very noisy (counts jumping up and down) then it can be helpful to 
calculate a moving average. In the example below, for each week we calculate the 
average number of cases from the four previous weeks. This smooths the data, to 
make it more interpretable. In our case this does not really add much, so we will
stick to the interpolated data for further analyis. 
See the **moving averages** page for more detail. 

```{r moving_averages}

## create a moving average variable (deals with missings)
counts <- counts %>% 
     ## create the ma_4w variable 
     ## slide over each row of the case variable
     mutate(ma_4wk = slide_dbl(case, 
                               ## for each row calculate the name
                               ~ mean(.x, na.rm = TRUE),
                               ## use the four previous weeks
                               .before = 4))

## make a quick visualisation of the difference 
ggplot(counts, aes(x = epiweek)) + 
     geom_line(aes(y = case)) + 
     geom_line(aes(y = ma_4wk), colour = "red")

```


<!-- ======================================================= -->
### Periodicity {.tabset .tabset-fade .tabset-pills}

```{r periodogram}

## x is a dataset
## counts is variable with count data or rates within x 
## start_week is the first week in your dataset
## period is how many units in a year 
## output is whether you want return spectral periodogram or the peak weeks
  ## "periodogram" or "weeks"
periodogram <- function(x, 
                        counts, 
                        start_week = c(2002, 1), 
                        period = 52, 
                        output = "weeks") {
  

    ## make sure is not a tsibble, filter to project and only keep columns of interest
    prepare_data <- dplyr::as_tibble(x)
    # prepare_data <- prepare_data[prepare_data[[strata]] == j, ]
    prepare_data <- dplyr::select(prepare_data, {{counts}})
    
    ## create an intermediate "zoo" time series to be able to use with spec.pgram
    zoo_cases <- zoo::zooreg(prepare_data, 
                             start = start_week, frequency = period)
    
    ## get a spectral periodogram not using fast fourier transform 
    periodo <- spec.pgram(zoo_cases, fast = FALSE, plot = FALSE)
    
    ## return the peak weeks 
    periodo_weeks <- 1 / periodo$freq[order(-periodo$spec)] * period
    
    if (output == "weeks") {
      periodo_weeks
    } else {
      periodo
    }
    
}

## get spectral periodogram for extracting weeks with the highest frequencies 
## (checking of seasonality) 
periodo <- periodogram(counts, 
                       case_int, 
                       start_week = c(2002, 1),
                       output = "periodogram")

## pull spectrum and frequence in to a dataframe for plotting
periodo <- data.frame(periodo$freq, periodo$spec)

## plot a periodogram showing the most frequently occuring periodicity 
ggplot(data = periodo, 
                aes(x = 1/(periodo.freq/52),  y = log(periodo.spec))) + 
  geom_line() + 
  labs(x = "Period (Weeks)", y = "Log(density)")


## get a vector weeks in ascending order 
peak_weeks <- periodogram(counts, 
                          case_int, 
                          start_week = c(2002, 1), 
                          output = "weeks")

```

<span style="color: black;">**_NOTE:_** It is possible to use the above weeks to add them to sin and cosine terms, however we will use a function to generate these terms (see regression section below) </span>

<!-- ======================================================= -->
### Decomposition {.tabset .tabset-fade .tabset-pills}

Classical decomposition is used to break a time series down several parts, which
when taken together make up for the pattern you see. 
These different parts are the trend-cycle (the long-term direction of the data), 
the seasonality (repeating patterns) and the random (what is left after removing 
trend and season). 

```{r decomposition}

## decompose the counts dataset 
counts %>% 
  # using an additive classical decomposition model
  model(classical_decomposition(case_int, type = "additive")) %>% 
  ## extract the important information from the model
  components() %>% 
  ## generate a plot 
  autoplot()

```

<!-- ======================================================= -->
### Autocorrelation {.tabset .tabset-fade .tabset-pills}

Autocorrelation tells you about the relation between the counts of each week 
and the weeks before it (called lags). 
Using the `ACF` function, we can produce a plot which shows us a number of lines 
for the relation at different lags. Where the lag is 0 (x = 0), this line would 
always be 1 as it shows the relation between an observation and itself (not shown here). 
The first line shown here (x = 1) shows the relation between each observation 
and the observation before it (lag of 1), the second shows the relation between 
each observation and the observation before last (lag of 2) and so on until lag of
52 which shows the relation between each observation and the observation from 1 
year (52 weeks before). 
Using the `PACF` function (for partial autocorrelation shows the same type of relation 
but adjusted for all other weeks between. This is less informative for determining
periodicity. 

```{r autocorrelation}

## using the counts dataset
counts %>% 
  ## calculate autocorrelation using a full years worth of lags
  ACF(case_int, lag_max = 52) %>% 
  ## show a plot
  autoplot()

## using the counts data set 
counts %>% 
  ## calculate the partial autocorrelation using a full years worth of lags
  PACF(case_int, lag_max = 52) %>% 
  ## show a plot
  autoplot()

```

You can formally test the null hypothesis of independence in a time series (i.e. 
that it is not autocorrelated) using the Ljung-Box test (in the **stats** package). 
A significant p-value suggests that there is autocorrelation in the data.

```{r ljung_box}

## test for independance 
Box.test(counts$case_int, type = "Ljung-Box")

```


<!-- ======================================================= -->
## Fitting regressions {.tabset .tabset-fade .tabset-pills}

It is possible to fit a large number of different regressions to a time series, 
however, here we will demonstrate how to fit a negative binomial regression - as 
this is often the most appropriate for counts data in infectious diseases. 

<!-- ======================================================= -->
### Fourier terms {.tabset .tabset-fade .tabset-pills}

Fourier terms are the equivalent of sin and cosin curves. The difference is that 
these are fit based on finding the most appropriate combination of curves to explain
your data. 
If only fitting one fourier term, this would be the equivalent of fitting a sin 
and a cosin for your most frequently occurring lag seen in your periodogram (in our 
case 52 weeks). We use the `fourier` function from the **forecast** package. 
In the below code we assign using the `$`, as `fourier` returns two columns (one 
for sin one for cosin) and so these are added to the dataset as a list, called 
"fourier" - but this list can then be used as a normal variable in regression. 

```{r fourier}

## add in fourier terms using the epiweek and case_int variabless
counts$fourier <- select(counts, epiweek, case_int) %>% 
  fourier(K = 1)
```

<!-- ======================================================= -->
### Negative binomial {.tabset .tabset-fade .tabset-pills}

It is possible to fit regressions using base **stats** or **MASS**
functions (e.g. `lm`, `glm` and `glm.nb`). However we will be using those from 
the **trending** package, as this allows for calculating appropriate confidence
and prediction intervals (which are otherwise not available). 
The syntax is the same, and you specify an outcome variable then a tilda (~) 
and then add your various exposure variables of interest separated by a plus (+). 

The other difference is that we first define the model and then `fit` it to the 
data. This is useful because it allows for comparing multiple different models 
with the same syntax. 

<span style="color: darkgreen;">**_TIP:_** For fitting more complex models such 
as ARIMA or prophet, see the [**fable**](https://fable.tidyverts.org/index.html) package.</span>

```{r nb_reg, warning = FALSE}

## define the model you want to fit (negative binomial) 
model <- glm_nb_model(
  ## set number of cases as outcome of interest
  case_int ~
    ## use epiweek to account for the trend
    epiweek +
    ## use the furier terms to account for seasonality
    fourier)

## fit your model using the counts dataset
fitted_model <- fit(model, counts)

## calculate confidence intervals and prediction intervals 
observed <- predict(fitted_model)

## plot your regression 
ggplot(data = observed, aes(x = epiweek)) + 
  ## add in a line for the model estimate
  geom_line(aes(y = estimate),
            col = "Red") + 
  ## add in a band for the prediction intervals 
  geom_ribbon(aes(ymin = lower_pi, 
                  ymax = upper_pi), 
              alpha = 0.25) + 
  ## add in a line for your observed case counts
  geom_line(aes(y = case_int), 
            col = "black") + 
  ## make a traditional plot (with black axes and white background)
  theme_classic()


```

<!-- ======================================================= -->
### Residuals {.tabset .tabset-fade .tabset-pills}

To see how well our model fits the observed data we need to look at the residuals. 
The residuals are the difference between the observed counts and the counts 
estimated from the model. We could calculate this simply by using `case_int - estimate`, 
but the `residuals` function extracts this directly from the regression for us.

What we see from the below, is that we are not explaining all of the variation 
that we could with the model. It might be that we should fit more fourier terms, 
and address the amplitude. However for this example we will leave it as is. 
The plots show that our model does worse in the peaks and troughs (when counts are
at their highest and lowest) and that it might be more likely to underestimate 
the observed counts. 

```{r}

## calculate the residuals 
observed <- observed %>% 
  mutate(resid = residuals(fitted_model$fitted_model, type = "response"))

## are the residuals fairly constant over time (if not: outbreaks? change in practice?)
observed %>%
  ggplot(aes(x = epiweek, y = resid)) +
  geom_line() +
  geom_point() + 
  labs(x = "epiweek", y = "Residuals")

## is there autocorelation in the residuals (is there a pattern to the error?)  
observed %>% 
  as_tsibble(index = epiweek) %>% 
  ACF(resid, lag_max = 52) %>% 
  autoplot()

## are residuals normally distributed (are under or over estimating?)  
observed %>%
  ggplot(aes(x = resid)) +
  geom_histogram(binwidth = 100) +
  geom_rug() +
  labs(y = "count") 
  
## compare observed counts to their residuals 
  ## should also be no pattern 
observed %>%
  ggplot(aes(x = estimate, y = resid)) +
  geom_point() +
  labs(x = "Fitted", y = "Residuals")

## formally test autocorellation of the residuals
## H0 is that residuals are from a white-noise series (i.e. random)
## test for independence 
## if p value significant then non-random
Box.test(observed$resid, type = "Ljung-Box")

```

<!-- ======================================================= -->
## Relation of two time series {.tabset .tabset-fade .tabset-pills}

Here we look at using weather data (specifically the temperature) to explain 
campylobacter case counts. 

<!-- ======================================================= -->
### Merging datasets {.tabset .tabset-fade .tabset-pills}

We can join our datasets using the week variable. For more on merging see the 
handbook section on [joining].

```{r join}

## left join so that we only have the rows already existing in counts
## drop the date variable from temp_data (otherwise is duplicated)
counts <- left_join(counts, 
                    select(temp_data, -date),
                    by = "epiweek")

```

<!-- ======================================================= -->
### Descriptive analysis {.tabset .tabset-fade .tabset-pills}

First plot your data to see if there is any obvious relation. 
The plot below shows that there is a clear relation in the seasonality of the two
variables, and that temperature might peak a few weeks before the case number.
For more on pivoting data, see the handbook section on [cleaning data]. 

```{r basic_plot_bivar}

counts %>% 
  ## keep the variables we are interested 
  select(epiweek, case_int, t2m) %>% 
  ## change your data in to long format
  pivot_longer(
    ## use epiweek as your key
    !epiweek,
    ## move column names to the new "measure" column
    names_to = "measure", 
    ## move cell values to the new "values" column
    values_to = "value") %>% 
  ## create a plot with the dataset above
  ## plot epiweek on the x axis and values (counts/celsius) on the y 
  ggplot(aes(x = epiweek, y = value)) + 
    ## create a separate plot for temperate and case counts 
    ## let them set their own y-axes
    facet_grid(measure ~ ., scales = "free_y") +
    ## plot both as a line
    geom_line()

```

<!-- ======================================================= -->
### Lags and cross-correlation {.tabset .tabset-fade .tabset-pills}

To formally test which weeks are most highly related between cases and temperature. 
We can use the cross-correlation function (`CCF`) from the **feasts** package. 
You could also visualise (rather than using `arrange`) using the `autoplot` function. 

```{r cross_correlation}

counts %>% 
  ## calculate cross-correlation between interpolated counts and temperature
  CCF(case_int, t2m,
      ## set the maximum lag to be 52 weeks
      lag_max = 52, 
      ## return the correlation coefficient 
      type = "correlation") %>% 
  ## arange in decending order of the correlation coefficient 
  ## show the most associated lags
  arrange(-ccf) %>% 
  ## only show the top ten 
  slice_head(n = 10)

```

We see from this that a lag of 4 weeks is most highly correlated, 
so we make a lagged temperature variable to include in our regression. 

```{r lag_tempvar}

counts <- counts %>% 
  ## create a new variable for temperature lagged by four weeks
  mutate(t2m_lag4 = lag(t2m, n = 4))

```


<!-- ======================================================= -->
### Negative binomial with two variables {.tabset .tabset-fade .tabset-pills}

We fit a negative binomial regression as done previously. This time we add the 
temperature variable lagged by four weeks. 

```{r nb_reg_bivar, warning = FALSE}

## define the model you want to fit (negative binomial) 
model <- glm_nb_model(
  ## set number of cases as outcome of interest
  case_int ~
    ## use epiweek to account for the trend
    epiweek +
    ## use the furier terms to account for seasonality
    fourier + 
    ## use the temperature lagged by four weeks 
    t2m_lag4
    )

## fit your model using the counts dataset
fitted_model <- fit(model, counts)

## calculate confidence intervals and prediction intervals 
observed <- predict(fitted_model)

```


To investigate the individual terms, we can pull the original negative binomial
regression out of the **trending** format using `get_model` and pass this to the
**broom** package `tidy` function to retrieve exponentiated estimates and associated
confidence intervals. 
What this shows us is that lagged temperature, after controlling for trend and seasonality, 
is similar to the case counts (estimate ~ 1) and significantly associated. 
This suggests that it might be a good variable for use in predicting future case
numbers (as climate forecasts are readily available). 

```{r results_nb_reg_bivar}

fitted_model %>% 
  ## extract original negative binomial regression
  get_model() %>% 
  ## get a tidy dataframe of results
  tidy(exponentiate = TRUE, 
       conf.int = TRUE)
```

A quick visual inspection of the model shows that it might do a better job of 
estimating the observed case counts. 

```{r plot_nb_reg_bivar}

## plot your regression 
ggplot(data = observed, aes(x = epiweek)) + 
  ## add in a line for the model estimate
  geom_line(aes(y = estimate),
            col = "Red") + 
  ## add in a band for the prediction intervals 
  geom_ribbon(aes(ymin = lower_pi, 
                  ymax = upper_pi), 
              alpha = 0.25) + 
  ## add in a line for your observed case counts
  geom_line(aes(y = case_int), 
            col = "black") + 
  ## make a traditional plot (with black axes and white background)
  theme_classic()


```


### Residuals {.tabset .tabset-fade .tabset-pills}

We investigate the residuals again to see how well our model fits the observed data. 
The results and interpretation here are similar to those of the previous regression, 
so it may be more feasible to stick with the simpler model without temperature. 

```{r}

## calculate the residuals 
observed <- observed %>% 
  mutate(resid = case_int - estimate)

## are the residuals fairly constant over time (if not: outbreaks? change in practice?)
observed %>%
  ggplot(aes(x = epiweek, y = resid)) +
  geom_line() +
  geom_point() + 
  labs(x = "epiweek", y = "Residuals")

## is there autocorelation in the residuals (is there a pattern to the error?)  
observed %>% 
  as_tsibble(index = epiweek) %>% 
  ACF(resid, lag_max = 52) %>% 
  autoplot()

## are residuals normally distributed (are under or over estimating?)  
observed %>%
  ggplot(aes(x = resid)) +
  geom_histogram(binwidth = 100) +
  geom_rug() +
  labs(y = "count") 
  
## compare observed counts to their residuals 
  ## should also be no pattern 
observed %>%
  ggplot(aes(x = estimate, y = resid)) +
  geom_point() +
  labs(x = "Fitted", y = "Residuals")

## formally test autocorrelation of the residuals
## H0 is that residuals are from a white-noise series (i.e. random)
## test for independence 
## if p value significant then non-random
Box.test(observed$resid, type = "Ljung-Box")

```


<!-- ======================================================= -->
## Resources {.tabset .tabset-fade .tabset-pills}

This tab should stay with the name "Resources".
Links to other online tutorials or resources.





