
# Transmission chains {.tabset .tabset-fade}


<!-- ======================================================= -->
## Overview {.tabset .tabset-fade .tabset-pills}

The primary tool to handle, analyse and visualise transmission chains and contact
tracing data is the package **epicontacts**, developed by the folks at
RECON. Try out the interactive plot below by hovering over the nodes for more
information, dragging them to move them and clicking on them to highlight downstream cases.

```{r out.width=c('25%', '25%'), fig.show='hold', echo=F}

## install development version of epicontacts
if(
  !"epicontacts" %in% rownames(installed.packages()) |
  packageVersion("epicontacts") != "1.2.0"
) remotes::install_github("reconhub/epicontacts@timeline")

## install and load packages
pacman::p_load(tidyverse, epicontacts, magrittr, here)

## load linelist
linelist <- rio::import(here::here("data", "linelist_cleaned.rds")) %>%
  filter(!duplicated(case_id))

## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id,
    location = sample(c("Community", "Nosocomial"), n(), TRUE),
    duration = sample.int(10, n(), TRUE)
  ) %>%
  drop_na(from)

## generate epicontacts
epic <- epicontacts::make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)

## subset object
epic %<>% subset(
  node_attribute = list(date_onset = c(as.Date(c("2014-06-01", "2014-07-01"))))
) %>%
  thin("contacts")

## plot with date of onset as x-axis
plot(
  epic,
  x_axis = "date_onset",
  label = FALSE,
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  node_shape = "gender",
  shapes = c(f = "female", m = "male"),
  unlinked_pos = "bottom",
  date_labels = "%b %d %Y",
  node_size = 35,
  font_size = 20,
  arrow_size = 0.5,
  height = 800,
  width = 700,
  edge_linetype = "location",
  legend_width = 0.15,
  highlight_downstream = TRUE,
  selector = FALSE
)

```

<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}

### Packages and data 

First load the standard packages required for data import and manipulation. 
	
```{r transmission_chains_packages, eval = FALSE}
pacman::p_load(
   rio,          # File import
   here,         # File locator
   tidyverse,    # Data management + ggplot2 graphics
   remotes       # Package installation from github
)
```
	
You will require the development version of **epicontacts**, which can be
installed from github using the **remotes** package. You only need to run the code
below once, not every time you use the package.

```{r transmission_chains_epicontacts_install, eval = FALSE}
remotes::install_github("reconhub/epicontacts@timeline")
```

Next, import the standard, cleaned linelist for this analysis.

```{r transmission_chains_load_show, eval=F}
# import the cleaned linelist
linelist <- rio::import("linelist_cleaned.xlsx")
```

### Creating an epicontacts object

We then need to create an **epicontacts** object, which requires two types of
data:

* a linelist documenting cases where columns are variables and rows correspond to unique cases
* a list of edges defining links between cases on the basis of their unique IDs (these can be contacts,
  transmission events, etc.)

As we already have a linelist, we just need to create a list of edges between
cases, more specifically between their IDs. We can extract transmission links from the
linelist by linking the `infector` column with the `case_id` column. At this point we can also add "edge
properties", by which we mean any variable describing the link between the two
cases, not the cases themselves. For illustration, we will add a `location`
variable describing the location of the transmission event, and a duration
variable describing the duration of the contact in days.

In the code below, the **dplyr** function `transmute` is similar to `mutate`, except it only keeps
the columns we have specified within the function. The `drop_na` function will
filter out any rows where the specified columns have an `NA` value; in this
case, we only want to keep the rows where the infector is known.

```{r transmission_chains_create_contacts, eval=T}
## generate contacts
contacts <- linelist %>%
  transmute(
    infector = infector,
    case_id = case_id,
    location = sample(c("Community", "Nosocomial"), n(), TRUE),
    duration = sample.int(10, n(), TRUE)
  ) %>%
  drop_na(infector)
```

We can now create the **epicontacts** object using the `make_epicontacts`
function. We need to specify which column in the linelist points to the unique case
identifier, as well as which columns in the contacts point to the unique
identifiers of the cases involved in each link. These links are directional in
that infection is going _from_ the infector _to_ the case, so we need to specify
the `from` and `to` arguments accordingly. We therefore also set the `directed`
argument to `TRUE`, which will affect future operations.

```{r transmission_chains_create_epicontacts, eval=T}
## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts,
  id = "case_id",
  from = "infector",
  to = "case_id",
  directed = TRUE
)
```
Upon examining the **epicontacts** objects, we can see that the `case_id` column
in the linelist has been renamed to `id` and the `case_id` and `infector`
columns in the contacts have been renamed to `from` and `to`. This ensures
consistency in subsequent handling, visualisation and analysis operations.

```{r transmission_chains_view_epicontacts, eval=T}
## view epicontacts object
epic
```

<!-- ======================================================= -->
## Handling {.tabset .tabset-fade .tabset-pills}

### Subsetting

The `subset()` method for `epicontacts` objects allows for, among other things,
filtering of networks based on properties of the linelist ("node attributes") and the contacts
database ("edge attributes"). These values must be passed as named lists to the
respective argument. For example, in the code below we are keeping only the
male cases in the linelist that have an infection date between April and
July 2014 (dates are specified as ranges), and transmission links that occured
in the hospital.

```{r transmission_chains_subset_nodes, eval=T}
sub_attributes <- subset(
  epic,
  node_attribute = list(
    gender = "m",
    date_infection = as.Date(c("2014-04-01", "2014-07-01"))
  ), 
  edge_attribute = list(location = "Nosocomial")
)
sub_attributes
```

We can use the `thin` function to either filter the linelist to include cases
that are found in the contacts by setting the argument `what = "linelist"`, or
filter the contacts to include cases that are found in the linelist by setting
the  argument `what = "contacts"`. In the code below, we are further filtering the
epicontacts object to keep only the transmission links involving the male cases
infected between April and July which we had filtered for above. We can see that
only two known transmission links fit that specification.

```{r transmission_chains_thin, eval=T}
sub_attributes <- thin(sub_attributes, what = "contacts")
nrow(sub_attributes$contacts)
```

In addition to subsetting by node and edge attributes, networks can be pruned to
only include components that are connected to certain nodes. The `cluster_id`
argument takes a vector of case IDs and returns the linelist of individuals that
are linked, directly or indirectly, to those IDs. In the code below, we can see
that a total of 13 linelist cases are involved in the clusters containing
`2ae019` and `71577a`.

```{r}
sub_id <- subset(epic, cluster_id = c("2ae019","71577a"))
nrow(sub_id$linelist)
```

The `subset()` method for `epicontacts` objects also allows filtering by cluster
size using the `cs`, `cs_min` and `cs_max` arguments. In the code below, we are
keeping only the cases linked to clusters of 10 cases or larger, and can see that
271 linelist cases are involved in such clusters.
    
```{r}   
sub_cs <- subset(epic, cs_min = 10)
nrow(sub_cs$linelist)
```

### Accessing IDs

The `get_id()` function retrieves information on case IDs in the
dataset, and can be parameterized as follows:

- **linelist**: IDs in the line list data
- **contacts**: IDs in the contact dataset ("from" and "to" combined)
- **from**: IDs in the "from" column of contact datset
- **to** IDs in the "to" column of contact dataset
- **all**: IDs that appear anywhere in either dataset
- **common**: IDs that appear in both contacts dataset and line list
    
For example, what are the first ten IDs in the contacts dataset?
```{r transmission_chains_get_ids, eval=T}
contacts_ids <- get_id(epic, "contacts")
head(contacts_ids, n = 10)
```

How many IDs are found in both the linelist and the contacts?
```{r transmission_chains_get_both, eval=T}
length(get_id(epic, "common"))
```

<!-- ======================================================= -->
## Visualization {.tabset .tabset-fade .tabset-pills}

### Basic plotting

All visualisations of **epicontacts** objects are handled by the `plot`
function. We will first filter the **epicontacts** object to include only the
cases with onset dates in June 2014 using the `subset` function, and only
include the contacts linked to those cases using the `thin` function.
	
```{r transmission_chains_basic_plot_sub, eval=T}
## subset epicontacts object
sub <- epic %>%
  subset(
    node_attribute = list(date_onset = c(as.Date(c("2014-06-30", "2014-06-01"))))
  ) %>%
 thin("contacts")
```

We can then create the basic, interactive plot very simply as follows:

```{r transmission_chains_basic_plot, eval=T}
## plot epicontacts object
plot(
  sub,
  width = 700,
  height = 700
)
```

You can move the nodes around by dragging them, hover over them for more
information and click on them to highlight connected cases.

There are a large number of arguments to further modify this plot. We will cover
the main ones here, but check out the documentation via `?vis_epicontacts` (the
function called when using `plot` on an **epicontacts** object) to get a full
description of the function arguments.

#### Visualising node attributes

Node color, node shape and node size can be mapped to a given column in the linelist 
using the `node_color`, `node_shape` and `node_size` arguments. This is similar
to the `aes` syntax you may recognise from **ggplot2**. 

The specific colors, shapes and sizes of nodes can be specified as follows:

* **Colors** via the `col_pal` argument, either by providing a name list for manual
specification of each color as done below, or by providing a color palette
function such as `colorRampPalette(c("black", "red", "orange"))`, which would
provide a gradient of colours between the ones specified.

* **Shapes** by passing a named list to the `shapes` argument, specifying one shape
  for each unique element in the linelist column specified by the `node_shape`
  argument. See `codeawesome` for available shapes.

* **Size** by passing a size range of the nodes to the `size_range` argument.

Here an example, where color represents the outcome, shape the gender and size
the age:

```{r transmission_chains_node_attribute, eval=T}
plot(
  sub, 
  node_color = "outcome",
  node_shape = "gender",
  node_size = 'age',
  col_pal = c(Death = "firebrick", Recover = "green"),
  shapes = c(f = "female", m = "male"),
  size_range = c(40, 60),
  height = 700,
  width = 700
)
```

#### Visualising edge attributes

Edge color, width and linetype can be mapped to a given column in the contacts
dataframe using the `edge_color`, `edge_width` and `edge_linetype`
arguments. The specific colors and widths of the edges can be specified as follows:

* **Colors** via the `edge_col_pal` argument, in the same manner used for `col_pal`.

* **Widths** by passing a size range of the nodes to the `width_range` argument.

Here an example:

```{r transmission_chains_edge_attribute, eval=T}
plot(
  sub, 
  node_color = "outcome",
  node_shape = "gender",
  node_size = 'age',
  col_pal = c(Death = "firebrick", Recover = "green"),
  shapes = c(f = "female", m = "male"),
  size_range = c(40, 60),
  edge_color = 'location',
  edge_linetype = 'location',
  edge_width = 'duration',
  edge_col_pal = c(Community = "orange", Nosocomial = "purple"),
  width_range = c(1, 3),
  height = 700,
  width = 700
)
```

### Temporal axis

We can also visualise the network along a temporal axis by mapping the `x_axis`
argument to a column in the linelist. In the example below, the x-axis
represents the date of symptom onset. We have also specified the `arrow_size`
argument to ensure the arrows are not too large, and set `label = FALSE` to make
 the figure less cluttered.

```{r transmission_chains_x_axis, eval=T}
plot(
  sub,
  x_axis = "date_onset",
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  arrow_size = 0.5,
  node_size = 13,
  label = FALSE,
  height = 700,
  width = 700
)
```

There are a large number of additional arguments to futher specify how this
network is visualised along a temporal axis, which you can check out
via `?vis_temporal_interactive` (the function called when using `plot` on
an **epicontacts** object with `x_axis` specified). We'll go through some
below.

#### Specifying transmission tree shape

There are two main shapes that the transmission tree can assume, specified using
the `network_shape` argument. The first is a `branching` shape as shown above,
where a straight edge connects any two nodes. This is the most intuitive
representation, however can result in overlapping edges in a densely connected
network. The second shape is `rectangle`, which produces a tree resembling a
phylogeny. For example:

```{r transmission_chains_rectangle, eval=T}
plot(
  sub,
  x_axis = "date_onset",
  network_shape = "rectangle",
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  arrow_size = 0.5,
  node_size = 13,
  label = FALSE,
  height = 700,
  width = 700
)
```

Each case node can be assigned a unique vertical position by toggling the
`position_dodge` argument. The position of unconnected cases (i.e. with no
reported contacts) is specified using the `unlinked_pos` argument.

```{r transmission_chains_dodge, eval=T}
plot(
  sub,
  x_axis = "date_onset",
  network_shape = "rectangle",
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  position_dodge = TRUE,
  unlinked_pos = "bottom",
  arrow_size = 0.5,
  node_size = 13,
  label = FALSE,
  height = 700,
  width = 700
)
```

The position of the parent node relative to the children nodes can be
specified using the `parent_pos` argument. The default option is to place the
parent node in the middle, however it can be placed at the bottom (`parent_pos =
'bottom'`) or at the top (`parent_pos = 'top'`).

```{r transmission_chains_parent_pos, eval=T}
plot(
  sub,
  x_axis = "date_onset",
  network_shape = "rectangle",
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  parent_pos = "top",
  arrow_size = 0.5,
  node_size = 13,
  label = FALSE,
  height = 700,
  width = 700
)
```

<!-- ======================================================= -->
## Analysis {.tabset .tabset-fade .tabset-pills}

### Summarising

We can get an overview of some of the network properties using the
`summary` function.

```{r transmission_chains_summarise_epicontacts, eval=T}
## summarise epicontacts object
summary(epic)
```

For example, we can see that only 57% of contacts have both cases in the
linelist; this means that the we do not have linelist data on a significant
number of cases involved in these transmission chains.

### Pairwise characteristics

The `get_pairwise()` function allows processing of variable(s) in the line list
according to each pair in the contact dataset. For the following example, date
of onset of disease is extracted from the line list in order to compute the
difference between disease date of onset for each pair. The value that is
produced from this comparison represents the **serial interval (si)**.

```{r transmission_chains_pairwise, eval=T}
si <- get_pairwise(epic, "date_onset")   
summary(si)
tibble(si = si) %>%
  ggplot(aes(si)) +
  geom_histogram() +
  labs(
    x = "Serial interval",
    y = "Frequency"
  )
```

The `get_pairwise()` will interpret the class of the column being used for
comparison, and will adjust its method of comparing the values accordingly. For
numbers and dates (like the **si** example above), the function will subtract
the values. When applied to columns that are characters or categorical,
`get_pairwise()` will paste values together. Because the function also allows
for arbitrary processing (see "f" argument), these discrete combinations can be
easily tabulated and analyzed.
    
```{r transmission_chains_pairwise_2, eval=T}
head(get_pairwise(epic, "gender"), n = 10)
get_pairwise(epic, "gender", f = table)
fisher.test(get_pairwise(epic, "gender", f = table))
```

Here, we see a significant association between transmission links and gender.

### Identifying clusters

The `get_clusters()` function can be used for to identify connected components
in an `epicontacts` object. First, we use it to retrieve a `data.frame`
containing the cluster information:

```{r transmission_chains_cluster, eval=T}
clust <- get_clusters(epic, output = "data.frame")
table(clust$cluster_size)
ggplot(clust, aes(cluster_size)) +
  geom_bar() +
  labs(
    x = "Cluster size",
    y = "Frequency"
  )
```

Let us look at the largest clusters. For this, we add cluster information to the
`epicontacts` object and then subset it to keep only the largest clusters:

```{r transmission_chains_cluster_2, eval=T}
epic <- get_clusters(epic)
max_size <- max(epic$linelist$cluster_size)
plot(subset(epic, cs = max_size))
```

### Calculating degrees

The degree of a node corresponds to its number of edges or connections to other
nodes. `get_degree()` provides an easy method for calculating this value for
`epicontacts` networks. A high degree in this context indicates an individual
who was in contact with many others. The `type` argument indicates that we want
to count both the in-degree and out-degree, the `only_linelist` argument
indicates that we only want to calculate the degree for cases in the linelist.

```{r transmission_chains_degree, eval=T}
deg_both <- get_degree(epic, type = "both", only_linelist = TRUE)
```

Which individuals have the ten most contacts?

```{r}
head(sort(deg_both, decreasing = TRUE), 10)
```

What is the mean number of contacts?

```{r}
mean(deg_both)
```

<!-- ======================================================= -->
## Resources {.tabset .tabset-fade .tabset-pills}

The
[epicontacts page](https://www.repidemicsconsortium.org/epicontacts/index.html)
provides an overview of the package functions and includes some more in-depth
vignettes.

The [github page](http://github.com/reconhub/epicontacts) can be used to raise
issues and request features.
