
# Transmission Chains {.tabset .tabset-fade}


<!-- ======================================================= -->
## Overview {.tabset .tabset-fade .tabset-pills}

The primary tool to visualize and analyze transmission chains is the package **epicontacts**, developed by the folks at RECON.  

```{r out.width=c('25%', '25%'), fig.show='hold', echo=F}
library(epicontacts)
library(outbreaks)

links <- epicontacts::make_epicontacts(linelist = mers_korea_2015$linelist,
                                       contacts = mers_korea_2015$contacts, 
                                       directed = TRUE)
# plot without time
# plot(links,
#      selector = FALSE,
#      height = 700,
#      width = 700)

# plot with date of onset as x-axis
plot(sim,
     x_axis = 'onset',
     height = 700,
     width = 700,
)
```



<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}


<!-- ======================================================= -->
## Visualization {.tabset .tabset-fade .tabset-pills}

```{r}

links <- epicontacts::make_epicontacts(linelist = mers_korea_2015$linelist,
                                       contacts = mers_korea_2015$contacts, 
                                       directed = TRUE)
# plot without time
plot(links,
     selector = FALSE,
     height = 700,
     width = 700)

```

And in a transmission tree, with *date of onset on the x-axis*:  

Note: this currently requires installing a development version of **epicontacts** from github... @ttree 

```{r}

# plot with date of onset as x-axis
plot(sim,
     x_axis = 'onset',
     height = 700,
     width = 700,
)

```

<!-- ======================================================= -->
## Analysis {.tabset .tabset-fade .tabset-pills}

```{r}
summary(links)
```


<!-- ======================================================= -->
## Resources {.tabset .tabset-fade .tabset-pills}

This tab should stay with the name "Resources".
Links to other online tutorials or resources.




