
# Epicurves

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      warning = FALSE,
                      message = FALSE)
```

<br>

**Maintainer:** Neale Batra (neale.batra@gmail.com)

**Code contributors:** ...

**Data contributors:** *outbreaks* package

**Version:** 1.0.0 





<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->
## Introduction {.tabset .tabset-fade .tabset-pills}


<!-- ======================================================= -->
### How to read this page
  
**Please note:**  

* *This is an HTML report. It is not online, you are only using your web brower to view this local file.*  

* *This report is ***best viewed with Google Chrome***. Some functions may not work in other browsers.*  

* *Please click through all the tabs to see different sections of the report.*  

* *You can click "Code" tabs on the right to see code.*  




<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->
## Task: Epidemic Curves {.tabset .tabset-fade .tabset-pills}

* Overview of epidemic curves...  
* Online resources and standards  
* Common pitfalls or mistakes...
TODO



<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->
## Preparation {.tabset .tabset-fade .tabset-pills}


<!-- ======================================================= -->
### Methods

The data preparation involves the following steps, detailed in the following tabs:


* **Load packages**: installs and load the packages required for the scripts  

* **Load data**: imports datasets

* **Clean data**: this section contains *ad hoc* data cleaning, i.e. which is
  not used in other reports (otherwise cleaning should be done in a dedicated
  report); this section is also used to create new variables used in the
  analyses


<!-- ======================================================= -->
### Load packages

This code chunk shows the loading of packages required for the analyses. 

```{r load_packages, eval=F}
# Create vector of names of required packages:
required_packages <- c("here",         # File locator
                       "rio",          # File import
                       "aweek",        # working with dates
                       "rmarkdown",    # Write Word, HTML, and Powerpoint files
                       "dplyr",        # data manipulation
                       "incidence",    # an option for epicurves of linelist data
                       "lubridate",    # Manipulate dates    
                       "stringr",      # Search and manipulate character strings
                       "forcats",      # working with factors
                       "ggplot2",      # Produce plots and graphs
                       "knitr",        # Produce tables
                       "kableExtra",   # produce detailed tables
                       "RColorBrewer"  # Color palettes from colorbrewer2.org
                       )               ### close vector of required packages

# Check if package is installed, installs if necessary, loads package for current session
for (pkg in required_packages) {
  
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  # load packages for use in this current R session 
  library(pkg, character.only = TRUE)
}
```



<!-- ======================================================= -->
### Load data

We load two example datasets for use in this document:  

* Linelist of individual data from an H7N9 outbreak in 2013  
* Aggregated case counts of a fictional disease in 2019

If viewing in Google Chrome, you can access these datasets in Microsoft Excel by clicking HERE and HERE.  

The data are imported using import() from the *rio* package. See the *page on importing data* for various ways to import data. The import() function wraps around the here() function of the *here* package, which serves to locate the file in its folder.  

Each data set is displayed below as a table for viewing.

For most of this document, the *linelist dataset* will be used. The aggregated counts dataset will be used at the end.  


```{r load_data_1, eval=F}
# import the linelist into R
linelist <- rio::import(here::here("epicurve", "data", "fluH7N9_china_2013.csv"))


# import the counts data into R
count_data <- rio::import(here::here("epicurve", "data", "Disease_reports_aggregated.xlsx"))

```


