# (PART) Introduction {-}

# Introduction {.tabset .tabset-fade}


<!-- ======================================================= -->
## About this handbook

**Objective**  

**A free open-access digital R reference book catered to epidemiologists and public health practitioners that is usable offline and addresses common epidemiological tasks via clear text explanations, step-by-step instructions, and best practice R code examples**

Epis using R must often Google search and read dozens of forum pages to complete common data manipulation and visualization epi tasks. However, field epidemiologists often work in low internet-connectivity environments and have limited technical support. This handbook aims to fill this gap.  

  
**How to read this handbook:**  

* The is an HTML file which *can* be viewed offline, and is best viewed with Google Chrome.  

* Click the "clipboard" icon in the upper-right of each code chunk to copy it.    

* Package names are written in bold (e.g. **dplyr**) and functions are written like this: `mutate()`. Packages referenced either in text or within code like this: `dplyr::mutate()`  


**Version**  
The latest version of this handbook can be found at this [github repository](https://github.com/nsbatra/R_epi_handbook).  



<!-- ======================================================= -->
## Style and editorial decisions {#style}


* The handbook generally uses **tidyverse** R coding style. Read more [here](https://www.tidyverse.org/)


**Emphasis on readability and instruction**  

We chose to frequently write code on new lines, in order to offer more understandable comments. As a result, code that *could* be written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%                    # group the rows by 'name'
  slice_max(date, n = 1, with_ties = F) # if there's a tie (of date), take the first row
```

...is often written like this:  

```{r, eval=F}
obs %>% 
  group_by(name) %>%   # group the rows by 'name'
  slice_max(
    date,              # keep row per group with maximum date value 
    n = 1,             # keep only the single highest row 
    with_ties = F)     # if there's a tie (of date), take the first row
```


**Types of notes**

```
FOR EXAMPLE: This is a boxed example

```

<span style="color: black;">**_NOTE:_** This is a note</span>

<span style="color: darkgreen;">**_TIP:_** This is a tip.</span>

<span style="color: orange;">**_CAUTION:_** This is a cautionary note.</span>

<span style="color: red;">**_DANGER:_** This is a warning.</span>






**Table of package, function, and other editorial decisions**  

See below some brief notes on decisions we made and our rationale. If you disagree or want to offer a new tool, please to join/start a conversation on our Github page.  

Decision          |     Considered     |    Outcome        |    Brief rationale   
----------------- | --------------------|---------------|--------------------------------------------------    
Epiweeks | aweek, lubridate | lubridate | consistency, package maintenance prospects  
   



<!-- ======================================================= -->
## Datasets used {#datasets}

Here the datasets used in this handbook will be described and will be "downloadable" via link (the files will be stored within the HTML, so available offline as well)

* Linelist (...)  Linelist for the 2013 (first wave) H7N9 outbreak in China ([source](https://datadryad.org/stash/dataset/doi:10.5061/dryad.2g43n))
* Aggregated case counts (...)  
* GIS coordinates (...)  
* GIS shapefile (...)  
* modeling dataset? (...)



<!-- ======================================================= -->
## Contributors

**Editor-in-Chief:** Neale Batra (neale.batra@gmail.com)

**Editorial core team:**  ...  

**Authors:** ...

**Reviewers:** ...  

**Advisors**  ...  

**Data contributors:**  
  *outbreaks* package

Some of this material comes from the R4Epis website, which was also made by some of the same people...

RECON packages

Photo credits (logo): CDC Public Image gallery; R Graph Gallery

